---
title: "Detection of Coffee Adulteration Using Hyperspectral Imaging and Ensemble Machine Learning: MSC+SG+2D"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "sky"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: true
---

## Introduction

This project investigates the effectiveness of ensemble learning methods in distinguishing authentic coffee from adulterated samples using the full spectral data derived from hyperspectral imaging (HSI). By utilizing the entire spectral range, the study avoids selective wavelength reduction, instead leveraging the richness of the spectral dataset to ensure comprehensive model inputs.

The ensemble learning framework integrates predictions from diverse base models, including Random Forest (RF), Support Vector Machines (SVM), Artificial Neural Networks (ANN), Linear Discriminant Analysis (LDA), and k-Nearest Neighbors (kNN). Using the stacking approach, the ensemble capitalizes on the strengths of individual algorithms, enhancing robustness and classification accuracy across varying levels of coffee adulteration.

The ensemble's performance is rigorously evaluated through 10-fold repeated cross-validation, offering a reliable assessment while reducing the risk of overfitting. This comprehensive approach highlights the model’s generalizability and robustness in identifying adulteration levels across a range of concentrations (e.g., 1%, 5%, 10%, and higher).

The study provides valuable insights into the application of ensemble learning to hyperspectral data analysis, focusing on computational efficiency, sensitivity, and specificity. It underscores the potential of advanced machine learning techniques to leverage full-spectrum data for robust food authenticity and fraud detection.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis) 
  library(NeuralNetTools)
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(caretEnsemble)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(Boruta)
  library(ggthemes)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: msc_sg_2d spectra

hsi_msc_sg_2d <- read.csv("coffee_instant_msc_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_2d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_msc_sg_2d$binary_class <- as.factor(hsi_msc_sg_2d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_msc_sg_2d[,c(6:229)]
y <- hsi_msc_sg_2d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
# Create histogram of spectral intensities with improved formatting
hist(as.matrix(X), 
     col = "lightblue",
     border = "white",  
     main = "Distribution of Spectral Intensity Values",  
     xlab = "Spectral Intensity (AU)",  
     ylab = "Frequency",  
     las = 1,  
     cex.main = 1.2,  
     cex.lab = 1.1,  
     breaks = 30)  

# Add grid lines 
grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")

# Add box around plot 
box(lwd = 2)                                    
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_2d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("slateblue", "sienna")) +  # Specify your custom colors here
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +  # Ensure legend names are retained
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 10))
```

## **Balance the data using SMOTE**

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

# PreProcess by PCA, Centering and Scaling Data

```{r}

# Reduction dimensionality to eliminate collinearity, centering and scaling

# Imbalanced Data
pca_Preprocess<- preProcess(train_data, method = c("center", "scale", "pca"))
data_trainPreprocessed <- predict(pca_Preprocess, train_data)
data_testPreprocessed <- predict(pca_Preprocess, test_data)

#SMOTE data
pca_Preprocess_smote<- preProcess(smote_data, method = c("center", "scale", "pca"))
data_trainPreprocessed_smote <- predict(pca_Preprocess_smote, smote_data)
data_testPreprocessed_smote <- predict(pca_Preprocess_smote, test_data)
```

```{r}
# Check the dimensions to ensure consistency 
dim(data_trainPreprocessed)
dim(data_testPreprocessed)
dim(data_trainPreprocessed_smote)
dim(data_testPreprocessed_smote)
```


## Set Up Training Parameters

```{r}
# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# OneSE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```
```{r}
# We will train all the models: kNN, NNET, SVM, LDA and RF as base learners. Then the predictions from this model will be trained by RF as a  meta-learner before testing on an external validation set.

# Hyper-parameter Tuning: We will train the model with their own set of hyper parameter settings.

# LDA model: No hyper parameter

# k-NN
grid_knn <- expand.grid(.k = seq(3,30, by = 2))

# SVM. We will use a tuneLength of 10. This will automatically select optimal Cost (C) and sigma parameter for the RBF kernel 

# RF: We will also use a tuneLength of 10. This selects the optimal mtry/number of splits

# NNET model. Selects the optimal number of neurons and decay value  (regularization parameter)
grid_nnet <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

## Train the Models

```{r cache=TRUE}

# Unmodified data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_msc_sg_2d <- train(binary_class ~., data = data_trainPreprocessed,
                       method = 'lda',trControl = control, metric = 'Accuracy')

# 2. Train the KNN model
fit_knn_model_msc_sg_2d <- train(binary_class ~., data = data_trainPreprocessed,
                       method = 'knn',trControl = control, 
                       tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_msc_sg_2d <- train(binary_class ~., data = data_trainPreprocessed,
                       method = 'svmRadial',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_msc_sg_2d <- train(binary_class ~., data = data_trainPreprocessed,
                       method = 'rf',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_msc_sg_2d<-train(binary_class ~., data = data_trainPreprocessed, 
                      method = 'nnet', tuneGrid = grid_nnet,
                      trControl = control, metric = 'Accuracy',
                      trace = FALSE, MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# After training the models, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list = list(lda = fit_lda_model_msc_sg_2d, 
                   knn = fit_knn_model_msc_sg_2d,
                   svm = fit_svm_model_msc_sg_2d,
                   rf = fit_rf_model_msc_sg_2d,
                   nnet = fit_nnet_msc_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_msc_sg_2d <- caretStack(models_list, method = 'rf', metric = metric,
                                       trControl = control,tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```
```{r}
# Train the models on SMOTE balanced data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_smote_msc_sg_2d <- train(binary_class ~.,
                                       data = data_trainPreprocessed_smote,
                                       method = 'lda',trControl = control, 
                                       metric = 'Accuracy')

# 2. Train the kNN model
fit_knn_model_smote_msc_sg_2d <- train(binary_class ~., 
                                 data = data_trainPreprocessed_smote,
                                 method = 'knn',trControl = control, 
                                 tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_smote_msc_sg_2d <- train(binary_class ~., 
                                       data = data_trainPreprocessed_smote,
                                       method = 'svmRadial',trControl = control,
                                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_smote_msc_sg_2d <- train(binary_class ~., 
                                data = data_trainPreprocessed_smote,
                                method = 'rf',trControl = control,
                                tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_smote_msc_sg_2d<-train(binary_class ~., 
                          data = data_trainPreprocessed_smote,
                          method = 'nnet', tuneGrid = grid_nnet,
                          trControl = control, metric = 'Accuracy',
                          trace = FALSE, MaxNWts = 1000, maxit = 300) 
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```



```{r}
# Similarly, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list_smote = list(lda = fit_lda_model_smote_msc_sg_2d, 
                         knn = fit_knn_model_smote_msc_sg_2d,
                         svm = fit_svm_model_smote_msc_sg_2d,
                         rf = fit_rf_model_smote_msc_sg_2d,
                         nnet = fit_nnet_smote_msc_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_smote_msc_sg_2d <- caretStack(models_list_smote, 
                                             method = 'rf', 
                                             metric = metric,
                                             trControl = control,
                                             tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
print(fit_stack_ensemble_msc_sg_2d) # Model on imbalanced data
```

```{r}
print(fit_stack_ensemble_smote_msc_sg_2d) # SMOTE model
```


## Save the models for future use

```{r}
saveRDS(fit_stack_ensemble_msc_sg_2d, 
        "fit_stack_ensemble_msc_sg_2d.rds") # for unmodified data

saveRDS(fit_stack_ensemble_smote_msc_sg_2d, 
        "fit_stack_ensemble_smote_msc_sg_2d.rds") # SMOTE data
```

## Display the optimal parameters for the ensemble meta learner

```{r}
# Used 500 trees as a default
print(paste('The optimal parametrers for training the meta-learner rf model (on unmodified data) are mtry value of', 
            fit_stack_ensemble_msc_sg_2d$ens_model$bestTune)) # unmodified data 

print(paste('The optimal parametrers for training the meta-learner rf model (on data balanced by SMOTE) are  mtry value of', fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune)) # unmodified data  # SMOTE data 
```

## Display the cross-validation results in Tables

```{r warning=FALSE}
# Imbalanced data
kable(fit_stack_ensemble_msc_sg_2d$ens_model$results, 
      caption = "Cross-Validation Results on Imbalanced Data") # Imbalanced Data
```


```{r warning=FALSE}
# SMOTE data
kable(fit_stack_ensemble_smote_msc_sg_2d$ens_model$results,
      caption = "Cross-Validation Results on SMOTE Data") # Balanced Data
```

```{r}
# Select the best results row for unmodified data
selected_row <- fit_stack_ensemble_msc_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_msc_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)
```

```{r}
# Select the best results row for SMOTE Data
selected_row <- fit_stack_ensemble_smote_msc_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)

```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_stack_ensemble_msc_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_msc_sg_2d$ens_model$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_stack_ensemble_smote_msc_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_smote_msc_sg_2d$ens_model$pred$obs) # SMOTE data
```

Visualize Hyperparameter Tuning for Optimal Model Selection

```{r}

# Imbalanced Data

mtry_data <- fit_stack_ensemble_msc_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_msc_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

```{r}

# SMOTE Data

mtry_data <- fit_stack_ensemble_smote_msc_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```




## Test the model peformance on test data for unbalanced trained data

```{r}
# Make predictions with the ensemble model: Imbalanced Data

test_ensemble_unmodified <- predict(fit_stack_ensemble_msc_sg_2d,
                                newdata=data_testPreprocessed)

# Apply threshold to probabilities
class_predictions_unmodified <- ifelse(test_ensemble_unmodified[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_unmodified <- factor(class_predictions_unmodified, 
                                       levels = c("adulterated", "pure_arabica"))

reference_unmodified <- data_testPreprocessed$binary_class

# Get the confusion matrix

test_ensemble_unmodified_cfmatrix <- confusionMatrix(class_predictions_unmodified,
                                          reference_unmodified)
test_ensemble_unmodified_cfmatrix

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_unmodified_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Make predictions with the ensemble model: SMOTE Data

test_ensemble_smote <- predict(fit_stack_ensemble_smote_msc_sg_2d,
                                newdata=data_testPreprocessed_smote)

# Apply threshold to probabilities
class_predictions_smote <- ifelse(test_ensemble_smote[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_smote <- factor(class_predictions_smote, 
                                       levels = c("adulterated", "pure_arabica"))

reference_smote <- data_testPreprocessed_smote$binary_class

# Get the confusion matrix

test_ensemble_smote_cfmatrix <- confusionMatrix(class_predictions_smote,
                                          reference_smote)
test_ensemble_smote_cfmatrix

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_smote_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(ensemble_1 = fit_stack_ensemble_msc_sg_2d$ens_model, 
                          ensemble_smote = fit_stack_ensemble_smote_msc_sg_2d$ens_model))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert re-sampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results, col = 'red')

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_ensemble_unmodified_cfmatrix$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_smote <- as.data.frame(test_ensemble_smote_cfmatrix$table)
plt_smote$Reference <- factor(plt_smote$Reference, levels=rev(levels(plt_smote$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: ensemble Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_smote, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: ensemble SMOTE Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(data_testPreprocessed$binary_class,
                 as.numeric(class_predictions_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("ensemble"), col = 'blue', lwd = 3)
```

### Misclassifications

```{r}
## Imbalanced
observed <- y_test
predicted <- test_ensemble_unmodified
perc <- hsi_msc_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))

imbal_class$adulterated<- ifelse(imbal_class$adulterated >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

imbal_class$pure_arabica<- ifelse(imbal_class$pure_arabica >= 0.5, 
                             "pure_arabica",
                              "adulterated") 
                             

misclas_imbal <- imbal_class %>% dplyr::filter(observed!=adulterated & observed!=pure_arabica)
misclas_imbal

```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# SMOTE Data
observed <- y_test
predicted <- test_ensemble_smote
perc <- hsi_msc_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

bal_class <- as.data.frame(cbind(observed,predicted,perc))

bal_class$adulterated<- ifelse(bal_class$adulterated >= 0.5, 
                             "adulterated", 
                             "pure_arabica")
bal_class$pure_arabica<- ifelse(bal_class$pure_arabica >= 0.5, 
                             "pure_arabica",
                              "adulterated") 

misclas_smote <- bal_class %>% filter(observed!=adulterated & observed!=pure_arabica)
misclas_smote
```

```{r}
table(misclas_smote$perc_adulter)
```



