---
title: "Detection of Coffee Adulteration Using Hyperspectral Imaging and Linear Discriminant Analysis (LDA): Spectra Preprocessed by Standard Normal Variate, Savitzky Golay and Second Derivative (SNV+SG+2nd Derivative)"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: 'darkly'
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

Coffee adulteration, such as mixing Arabica with cheaper Robusta, compromises quality and authenticity. **Linear Discriminant Analysis (LDA)** is a classical statistical method that finds linear boundaries to maximize the separation between classes, making it well-suited for high-dimensional spectroscopic data. This study evaluates the capability of LDA to accurately classify pure and adulterated coffee, demonstrating its potential as a reliable tool for detecting and quantifying coffee adulteration.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(e1071) 
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(gridExtra)
  library(pROC)
  library(knitr)
  library(kernlab)}))
```

## Load and inspect data: snv_sg_2d spectra

```{r}
hsi_snv_sg_2d <- read.csv("coffee_ground_snv_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_snv_sg_2d)

# Check a few column names
col_check <- colnames(hsi_snv_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_snv_sg_2d$binary_class <- as.factor(hsi_snv_sg_2d$binary_class)
```

```{r}
# Get the X and y variables
X <- hsi_snv_sg_2d[,c(6:229)]
y <- hsi_snv_sg_2d[,2]
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue',
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()                                    
```

## Split samples into training and test sets

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model
sample<-hsi_snv_sg_2d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)

# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Chek composition of the y variable for both train and test sets
table(train_data$binary_class)
table(test_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Balance the data using SMOTE

```{r}
# Due to data imbalance, we will use SMOTE to balance the data set

library(DMwR)

# Combine the dataset
smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)

# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

# Apply PCA, Center and Scale Data

```{r}
#Reduction dimensionality with PCA to mitigate multicollinearity

pca_Preprocess<- preProcess(train_data, method = c("center", "scale", "pca"))
data_trainPreprocessed <- predict(pca_Preprocess, train_data)
data_testPreprocessed <- predict(pca_Preprocess, test_data)

#SMOTE data
pca_Preprocess_smote<- preProcess(smote_data, method = c("center", "scale", "pca"))
data_trainPreprocessed_smote <- predict(pca_Preprocess_smote, smote_data)
data_testPreprocessed_smote <- predict(pca_Preprocess_smote, test_data)
```

## Set Up Training Parameters

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```


## Train the Models

```{r cache=TRUE}
# Set seed for reproducibility
set.seed(123)
# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_lda_snv_sg_2d<-train(y = data_trainPreprocessed[,1], x = data_trainPreprocessed[,-1], 
                      method = 'lda', trControl = control, metric = 'Accuracy') #unmodified data
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

print(model_training_time) # Time taken to train the model
```

```{r}
print(fit_lda_snv_sg_2d)
```

```{r cache=TRUE}

set.seed(123) # Reproducibility

# Start_time
start_time<-Sys.time()

# Fit the model
fit_lda_smote_snv_sg_2d<-train(y = data_trainPreprocessed_smote[,1], x = data_trainPreprocessed_smote[,-1],
                            method ='lda', trControl = control, metric = 'Accuracy')# SMOTE data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time

print(model_training_time)
```

```{r}
print(fit_lda_smote_snv_sg_2d)
```

## Save the models for future use

```{r}
saveRDS(fit_lda_snv_sg_2d, "fit_lda_snv_sg_2d.rds")
saveRDS(fit_lda_smote_snv_sg_2d, "fit_lda_smote_snv_sg_2d.rds")
```


## Display the cross-validation results in Tables


```{r}
kable(fit_lda_snv_sg_2d$results, caption = 'LDA Cross-Validation Results (Imbalanced Data)')
```

```{r}
kable(fit_lda_smote_snv_sg_2d$results, caption = 'LDA Cross-Validation Results (SMOTE Data)')
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_lda_snv_sg_2d$pred$pred,
                fit_lda_snv_sg_2d$pred$obs) # unmodified data
```


```{r}
confusionMatrix(fit_lda_smote_snv_sg_2d$pred$pred,fit_lda_smote_snv_sg_2d$pred$obs) # SMOTE data
```

## LDA Plot for Unmodified Data

```{r}
# Project training data onto LDA components # unmodified Data

lda_projection <- predict(fit_lda_snv_sg_2d$finalModel, 
                          data_trainPreprocessed[,-1])

# Convert LDA results to a data frame
lda_data <- data.frame(
  Discriminant1 = lda_projection$x[, 1],
  Class = data_trainPreprocessed[, 1])

# Set the color and shape
group_colors <- c("red", "blue")  
group_shapes <- c(16, 17)  

# Plot the data
plot(lda_data$Discriminant1, 
     col = group_colors[as.numeric(data_trainPreprocessed[, 1])],
     pch = group_shapes[as.numeric(data_trainPreprocessed[, 1])],
     xlab = "Class",
     ylab = "Discriminant Function 1",
     main = "LDA Projection SMOTE Data", cex.main = 1.0,
     cex.axis = 0.8, cex.lab = 0.8, xaxt = "n")
# Add a legend
legend("bottomleft", legend = levels(data_trainPreprocessed[, 1]),
       col = group_colors, pch = group_shapes, title = "Classes", cex = 0.6)
```

## LDA Plot SMOTE Data

```{r}
# Project training data onto LDA components # SMOTE Data

lda_projection <- predict(fit_lda_smote_snv_sg_2d$finalModel, 
                          data_trainPreprocessed_smote[,-1])
head(lda_data)

# Convert LDA results to a data frame
lda_data <- data.frame(
  Discriminant1 = lda_projection$x[, 1],
  Class = data_trainPreprocessed_smote[, 1])

rownames(lda_data) <- NULL
# Set the color and shape
group_colors <- c("red", "blue")  
group_shapes <- c(16, 17)  

# Plot the data
plot(lda_data$Discriminant1, 
     col = group_colors[as.numeric(data_trainPreprocessed_smote[, 1])],
     pch = group_shapes[as.numeric(data_trainPreprocessed_smote[, 1])],
     xlab = "Class",
     ylab = "Discriminant Function 1",
     main = "LDA Projection", cex.main = 1.0,
     cex.axis = 0.8, cex.lab = 0.8, xaxt = "n")
# Add a legend
legend("bottomright", legend = levels(data_trainPreprocessed_smote[, 1]),
       col = group_colors, pch = group_shapes, title = "Classes", cex = 0.8)
```




## Test the model performance on test data for unbalanced trained data

```{r}
# Predict on the test set 

test_lda_unmodified <- predict(fit_lda_snv_sg_2d,newdata=data_testPreprocessed)

# Get the confusion matrix
test_lda_unmodified_CMtrx <- confusionMatrix(test_lda_unmodified,
                                             data_testPreprocessed$binary_class)
test_lda_unmodified_CMtrx

knitr::kable(test_lda_unmodified_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_lda_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_lda <- test_lda_unmodified_CMtrx$table
TP <- cf_table_lda[1,1]
TN <- cf_table_lda[2,2] 
FP <- cf_table_lda[2,1] 
FN <- cf_table_lda[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Predict on the test set 

test_lda_smote <- predict(fit_lda_smote_snv_sg_2d,newdata=data_testPreprocessed_smote)

# Get the confusion matrix
test_lda_smote_CMtrx <- confusionMatrix(test_lda_smote,
                                             data_testPreprocessed_smote$binary_class)
test_lda_smote_CMtrx

knitr::kable(test_lda_smote_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_lda_smote_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_lda_smote <- test_lda_smote_CMtrx$table
TP <- cf_table_lda_smote[1,1]
TN <- cf_table_lda_smote[2,2] 
FP <- cf_table_lda_smote[2,1] 
FN <- cf_table_lda_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(lda = fit_lda_snv_sg_2d, lda_smote = fit_lda_smote_snv_sg_2d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
p1<-ggplot(data = as.data.frame(test_lda_unmodified_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix lda Unbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(data = as.data.frame(test_lda_smote_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix lda with SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(data_testPreprocessed_smote$binary_class,
                 as.numeric(test_lda_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("LDA"), col = 'blue', lwd = 3)
```

#### Check misclassification 

```{r}

```


## Conclusion
The LDA shows good performance. The models excel at accurately identifying pure Arabica coffee as observed from the high specificity.