---
title: "Detection of Coffee Adulteration Using Hyperspectral Imaging and Artificial Neural Networks: Spectra Preprocessed by Multiplicative Scatter Correction, Savitzky Golay and First Derivative (MSC+SG+1st Der)"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: false
---

## Introduction

This project utilizes Artificial Neural Networks (ANN) to evaluate the ability to discriminate between authentic and adulterated coffee samples based on spectral data. ANN, with its capacity to model complex and non-linear relationships, is well-suited for high-dimensional datasets such as those derived from spectroscopic analyses. The study aims to determine if the model can effectively distinguish pure coffee from adulterated samples across varying levels of adulteration. By assessing its discriminatory power, this approach seeks to validate ANN's potential as a reliable tool for authenticity verification and quality control in the coffee industry.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class) 
  library(randomForest)
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE)
  library(ggthemes)
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(e1071) 
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(gridExtra)
  library(pROC)
  library(NeuralNetTools)
  library(kernlab)}))
```

## Load and inspect data: msc_sg_1d spectra

```{r}
hsi_msc_sg_1d <- read.csv("coffee_ground_msc_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_1d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_1d[,c(1:6)])
col_check
# The spectral values start from column # 6 -------> 229

# Convert the class to a factor
hsi_msc_sg_1d$binary_class <- as.factor(hsi_msc_sg_1d$binary_class)
```

```{r}
# Get the X and y variables
X <- hsi_msc_sg_1d[,c(6:229)]
y <- hsi_msc_sg_1d[,2]
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue',
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()    
```

## Split samples into training and test sets

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_1d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)

# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Chek composition of the y variable for both train and test sets
table(train_data$binary_class)
table(test_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Balance the data using SMOTE

```{r}
# Due to data imbalance, we will use SMOTE to balance the data set

warning = FALSE  
  suppressWarnings(suppressMessages({
library(DMwR)  
}))


# Combine the dataset
smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)

# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

# Reduction Dimension (PCA), Centering and Scaling Data

```{r}
# Reduction dimensionality to eliminate collinearity, centering and scaling

# Imabalanced Data
pca_Preprocess<- preProcess(train_data, method = c("center", "scale", "pca"))
data_trainPreprocessed <- predict(pca_Preprocess, train_data)
data_testPreprocessed <- predict(pca_Preprocess, test_data)

#SMOTE data
pca_Preprocess_smote<- preProcess(smote_data, method = c("center", "scale", "pca"))
data_trainPreprocessed_smote <- predict(pca_Preprocess_smote, smote_data)
data_testPreprocessed_smote <- predict(pca_Preprocess_smote, test_data)
```

## Set Up Training Parameters

```{r}
# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# OneSE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# Hyperparameter Tuning
# Set tune grid to find optimal value of size and decay on cross-validation
# Size is the number of number of units (neurons) in the hidden layer of the neural network
# Decay is the regularization parameter; adds penalty to the loss function

grid <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

## Train the Models

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_nnet_msc_sg_1d<-train(binary_class~., data = data_trainPreprocessed, 
                      method = 'nnet', tuneGrid = grid,
                      trControl = control, metric = 'Accuracy',
                      trace = FALSE, MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)

print(fit_nnet_msc_sg_1d)
```

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

fit_nnet_smote_msc_sg_1d<-train(binary_class~., data = data_trainPreprocessed_smote,
                            method ='nnet', tuneGrid = grid,
                            trControl = control, metric = 'Accuracy',
                            trace = FALSE, MaxNWts = 1000, maxit = 300)# SMOTE data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)

print(fit_nnet_smote_msc_sg_1d)
```

## Save the models for future use

```{r}
saveRDS(fit_nnet_msc_sg_1d, "fit_nnet_msc_sg_1d.rds")
saveRDS(fit_nnet_smote_msc_sg_1d, "fit_nnet_smote_msc_sg_1d.rds")
```

## Display the cross-validation results

```{r}
print(paste('The optimal parametrers for training the nnet model are', round(as.numeric(fit_nnet_msc_sg_1d$bestTune$size),2),'neurons','and decay value of',fit_nnet_msc_sg_1d$bestTune$decay)) # unmodified data 

print(paste('The optimal parametrers for training the nnet model are', round(as.numeric(fit_nnet_smote_msc_sg_1d$bestTune$size),2),'neurons','and decay value of',fit_nnet_smote_msc_sg_1d$bestTune$decay)) # SMOTE data 
```

```{r}
# Imbalanced data
knitr::kable(fit_nnet_msc_sg_1d$results)
```

```{r}
# Select the best results row 
selected_row <- fit_nnet_msc_sg_1d$results %>% 
  filter(size==fit_nnet_msc_sg_1d$bestTune$size & 
           decay ==fit_nnet_msc_sg_1d$bestTune$decay)


knitr::kable(selected_row)
```

```{r}
# SMOTE Data
knitr::kable(fit_nnet_smote_msc_sg_1d$results)

```

```{r}
# Select the best results row 
selected_row_smote <- fit_nnet_smote_msc_sg_1d$results %>% 
  filter(size==fit_nnet_smote_msc_sg_1d$bestTune$size & 
           decay ==fit_nnet_msc_sg_1d$bestTune$decay)

knitr::kable(selected_row_smote)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_nnet_msc_sg_1d$pred$pred,fit_nnet_msc_sg_1d$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_nnet_smote_msc_sg_1d$pred$pred,fit_nnet_smote_msc_sg_1d$pred$obs) # SMOTE data
```

Visualize Hyperparameter Tuning for Optimal Model Selection

```{r warning=FALSE}
# Imbalanced Data

# Get the best accuracy based on best size and decay
best_size <- fit_nnet_msc_sg_1d$bestTune$size
best_decay <- fit_nnet_msc_sg_1d$bestTune$decay
best_accuracy <- fit_nnet_msc_sg_1d$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_msc_sg_1d, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

```{r warning=FALSE}
# SMOTE data

# Get the best accuracy based on best size and decay
best_size <- fit_nnet_smote_msc_sg_1d$bestTune$size
best_decay <- fit_nnet_smote_msc_sg_1d$bestTune$decay
best_accuracy <- fit_nnet_smote_msc_sg_1d$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_smote_msc_sg_1d, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

## Visualize the NNET Model Architecture

```{r}
# Unmodified data
plotnet(fit_nnet_msc_sg_1d$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
```

```{r}
#SMOTE data
plotnet(fit_nnet_smote_msc_sg_1d$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
```

## Test the model peformance on test data for unbalanced trained data

```{r}
# Predict on the test set 
set.seed(123)
test_nnet_unmodified <- predict(fit_nnet_msc_sg_1d,newdata=data_testPreprocessed)

# Get the confusion matrix
test_nnet_unmodified_CMtrx <- confusionMatrix(test_nnet_unmodified,
                                             data_testPreprocessed$binary_class)
test_nnet_unmodified_CMtrx

knitr::kable(test_nnet_unmodified_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_nnet_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_nnet <- test_nnet_unmodified_CMtrx$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model pennetormance on test data using SMOTE trained model

```{r}
# Predict on the test set 
set.seed(123)
test_nnet_smote <- predict(fit_nnet_smote_msc_sg_1d,newdata=data_testPreprocessed_smote)

# Get the confusion matrix
test_nnet_smote_CMtrx <- confusionMatrix(test_nnet_smote,
                                             data_testPreprocessed_smote$binary_class)
test_nnet_smote_CMtrx

knitr::kable(test_nnet_smote_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_nnet_smote_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_nnet_smote <- test_nnet_smote_CMtrx$table
TP <- cf_table_nnet_smote[1,1]
TN <- cf_table_nnet_smote[2,2] 
FP <- cf_table_nnet_smote[2,1] 
FN <- cf_table_nnet_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model pennetormance
results <- resamples(list(nnet = fit_nnet_msc_sg_1d, nnet_smote = fit_nnet_smote_msc_sg_1d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Pennetormance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Dotplot for accuracy comparison
dotplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
p1<-ggplot(data = as.data.frame(test_nnet_unmodified_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix nnet Unbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(data = as.data.frame(test_nnet_smote_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "azure1", high = "turquoise", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix nnet with SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}

#SMOTE data
# Generate the ROC curve
roc_curve <- roc(data_testPreprocessed_smote$binary_class,
                 as.numeric(test_nnet_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')


# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("nnet"), col = 'blue', lwd = 3)
```

## Conclusion

-   NNET model achieve exceptional performance as observed from the ROC curve.
