---
title: "Cost Sensitive Learning: SVM Algorithm and Hyperspectral Imaging with SNV+SG+1st Derivative Spectra: Instant Coffee"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "league"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
  output_file: "output.html" 
  keep: always
editor: 
  markdown: 
    wrap: 72
---

## Introduction

In imbalanced classification problems, such as detecting pure Arabica
coffee (minority class) against adulterated variants (majority class),
traditional models may bias predictions toward the majority class,
resulting in poor performance on the minority class. To address this
issue, we implemented cost-sensitive learning using **SVM** with **class weights**.

**Cost-sensitive learning** assigns a higher penalty to
misclassifications of the minority class during training. This
encourages the model to pay more attention to underrepresented samples,
in this case, **pure Arabica**. This approach increases the contribution
of Arabica samples to the model’s optimization process, enhancing its
ability to correctly classify true Arabica instances.

### Why SVM?

Support Vector Machine (SVM) classification is a widely used machine learning algorithm well-suited for high-dimensional and imbalanced data. It supports class weighting, making it highly effective for cost-sensitive classification tasks. When misclassification penalties are properly adjusted, SVM offers a robust trade-off between:

🔍 Simplicity – straightforward to implement and tune with kernels and regularization
🧠 Margin-Based Decision Logic – focuses on maximizing class separation via support vectors
⚡ Performance – excels in handling nonlinear boundaries and complex, high-dimensional feature spaces

These strengths make SVM classification an excellent choice for applications where minority class detection is critical — such as fraud detection or authenticity verification in imbalanced datasets

## Setup and Load Packages

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(Boruta)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: snv_sg_1d spectra

hsi_snv_sg_1d <- read.csv("coffee_instant_snv_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_snv_sg_1d)

# Check a few column names
col_check <- colnames(hsi_snv_sg_1d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_snv_sg_1d$binary_class <- as.factor(hsi_snv_sg_1d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_snv_sg_1d[,c(6:229)]
y <- hsi_snv_sg_1d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any
    data leakage that could inflate model performance and make the model
    less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_snv_sg_1d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data



p0 <- train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
p0
```

-   There two classes are highly imbalanced!

## Perform Principal Component Analysis (PCA)

```{r}
pca_Preprocess <- preProcess(train_data, method = c("pca","center", "scale"))
train_mergedPreprocessed <- predict(pca_Preprocess, X_train) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,X_test) # apply on test data
```

```{r}
colnames(test_data)
```

```{r}
# Save the PCA object for model deploymet
saveRDS(pca_Preprocess,'pca_Preprocess.rds')
```

## Create a new data frame for training the model

```{r}
train_data_pca <- cbind(train_mergedPreprocessed, binary_class = y_train)
test_data_pca <- cbind(test_mergedPreprocessed, binary_class = y_test)
```

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# One SE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
# Set the metric as accuracy 
metric<-"Accuracy"
```

## Compute Class Weights

```{r}
class_counts <- table(train_data_pca$binary_class)
class_weights <- 1 / class_counts
sample_weights <- class_weights[train_data_pca$binary_class]
```

# Train svm Model

```{r}
# Imbalanced data

#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()
fit_svm_snv_sg_1d <- train(binary_class ~., 
                         data = train_data_pca,
                         method = 'svmRadial', 
                         trControl = control,tuneLength =10, 
                         metric = metric) 

#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# Print cross-validation results
print(fit_svm_snv_sg_1d)
```

```{r}
# Class Weighted LR Model

#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()
fit_svm_class_weight_snv_sg_1d <- train(binary_class ~., 
                         data = train_data_pca, method ='svmRadial',
                         trControl = control,tuneLength=10,
                         metric = metric, weights = sample_weights) 

#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# Print the cross-validation model
print(fit_svm_class_weight_snv_sg_1d)
```

## Save the models for future use

```{r}
saveRDS(fit_svm_snv_sg_1d,'fit_svm_svm_snv_sg_1d.rds')
saveRDS(fit_svm_class_weight_snv_sg_1d,"fit_svm__cost_sensitive.rds")
```

## Display Cross-Validation Results

```{r}
# Imbalanced data 
kable(fit_svm_snv_sg_1d$results)
```

```{r}
kable(fit_svm_class_weight_snv_sg_1d$results)
```

## Confusion Matrix for the training data 

```{r}
# Imbalanced data
confusionMatrix(fit_svm_snv_sg_1d$pred$pred,fit_svm_snv_sg_1d$pred$obs) 
```

```{r}
# Class-Weight svm ression 
confusionMatrix(fit_svm_class_weight_snv_sg_1d$pred$pred,fit_svm_class_weight_snv_sg_1d$pred$obs)
```

## Test the model performance on the test data for imbalanced data

```{r}
# Predict on the test set 
set.seed(123)
predictions_svm_imbal <- predict(fit_svm_snv_sg_1d,newdata=test_data_pca)

# Get the confusion matrix
test_svm_CMtrx <- confusionMatrix(predictions_svm_imbal,
                                             test_data_pca$binary_class)
test_svm_CMtrx
```

```{r}
# Confusion Table Imbalanced Data 
kable(test_svm_CMtrx$byClass)
```

```{r}
# View the results as knitr Table
kable(test_svm_CMtrx$table)
```

```{r}
# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm <- test_svm_CMtrx$table
TP <- cf_table_svm[1,1]
TN <- cf_table_svm[2,2] 
FP <- cf_table_svm[2,1] 
FN <- cf_table_svm[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on the test data based on cost-sensitive learning

```{r}
# Predict on the test set 
set.seed(123)
predictions_svm_class_wt <- predict(fit_svm_class_weight_snv_sg_1d,newdata=test_data_pca)

# Get the confusion matrix
test_svm_class_wt_CMtrx <- confusionMatrix(predictions_svm_class_wt,
                                             test_data_pca$binary_class)
test_svm_class_wt_CMtrx
```

```{r}
# Confusion Table based on class weights
kable(test_svm_class_wt_CMtrx$byClass)
```

```{r}
# View the results as knitr Table
kable(test_svm_class_wt_CMtrx$table)
```

```{r}
# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm_class_wt <- test_svm_class_wt_CMtrx$table
TP <- cf_table_svm_class_wt[1,1]
TN <- cf_table_svm_class_wt[2,2] 
FP <- cf_table_svm_class_wt[2,1] 
FN <- cf_table_svm_class_wt[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

```{r}
# Compare model performance
results <- resamples(list(svm_imbalanced = fit_svm_snv_sg_1d, 
                          svm_class_weight = fit_svm_class_weight_snv_sg_1d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")
```

```{r}
bwplot(results)
```

```{r}
# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics
```

## **Visualize the confusion matrix tables**

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_svm_CMtrx$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_class_weight <- as.data.frame(test_svm_class_wt_CMtrx$table)
plt_class_weight$Reference <- factor(plt_class_weight$Reference, 
                                     levels=rev(levels(plt_class_weight$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black",linewidth = 1) +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_class_weight, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black", linewidth = 1) +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Class-Weight Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8, color = 'black'),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}

# Generate the ROC curve
roc_curve <- roc(test_data_pca$binary_class,
                 as.numeric(predictions_svm_class_wt),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')

# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("SVM"), col = 'blue', lwd = 3)
```

## Missclassifications 

```{r}
## Imbalanced
observed <- y_test
predicted <- predictions_svm_imbal
perc <- hsi_snv_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_imbal <- imbal_class %>% filter(observed!=predicted)
misclas_imbal
```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# Class Weight svm ression 
observed <- y_test
predicted <- predictions_svm_class_wt
perc <- hsi_snv_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

class_wt <- as.data.frame(cbind(observed,predicted,perc))
misclas_class_wt <- class_wt%>% filter(observed!=predicted)
misclas_class_wt
```

```{r}
table(misclas_class_wt$perc_adulter)
```

