---
title: "Optimized Feature Selection Using Support Vector Machines with Recursive Feature Elimination and Genetic Algorithm for Detecting Coffee Adulteration via Hyperspectral Imaging: A Case for MSC+SG+1st Derivative Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "sky"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: false
---

## Introduction

Coffee adulteration, such as blending premium Arabica with lower-cost Robusta, undermines both quality and authenticity, posing significant challenges for consumers and regulators. High-dimensional spectroscopic data offers a wealth of information for detecting such adulteration, but the abundance of variables requires robust feature selection techniques to identify the most relevant wavelengths. Recursive Feature Elimination (RFE) and Genetic Algorithms (GA) are powerful tools for optimizing feature selection in spectroscopic datasets. This study integrates RFE and GA to enhance the performance of machine learning models in classifying pure and adulterated coffee. By leveraging these approaches, we aim to demonstrate an efficient and reliable method for detecting and quantifying coffee adulteration.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: msc_sg_1d spectra

hsi_msc_sg_1d <- read.csv("coffee_instant_msc_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_1d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_1d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_msc_sg_1d$binary_class <- as.factor(hsi_msc_sg_1d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_msc_sg_1d[,c(6:229)]
y <- hsi_msc_sg_1d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
hist(as.matrix(X), 
     col = "blue",
     border = "white",  
     main = "Distribution of Spectral Intensity Values",  
     xlab = "Spectral Intensity (AU)",  
     ylab = "Frequency",  
     las = 1,  
     cex.main = 1.2,  
     cex.lab = 1.1,  
     breaks = 30)  

# Add grid lines 
grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")

# Add box around plot 
box(lwd = 2)
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_1d$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Chek composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## **Balance the data using SMOTE**

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Define the Fitness Function for GA

-   This function evaluates subsets of features selected by GA using svme with a Random Forest (svm) model.

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}

# Fitness Funtion
fitness_ga_svme <- function(subset) {
  # Identify selected features based on the binary subset
  selected_features <- which(subset == 1)
  
  # Penalize empty subsets to avoid errors
  if (length(selected_features) == 0) return(Inf)
  
  # Subset the training feature matrix to selected features
  X_selected <- X_train[, selected_features]
  
  # Define control for svme
  control <- svmeControl(functions = svmFuncs,  # Use Random Forest for svme
                        method = "cv",       # Cross-validation
                        number = 10)         # 10-fold CV
  
  # Apply svme to the selected subset of features
  svme_result <- svme(X_selected, y_train, svmeControl = control)
  
  # Return the negative accuracy as the fitness score (to maximize accuracy)
  return(-svme_result$results$Accuracy[which.max(svme_result$results$Accuracy)])
}
```

## Run the Genetic Algorithm

```{r}

# Set seed for reproducibility 
set.seed(123)

num_features <- ncol(X_train) # Number of features
dataset_name <- "msc_sg_1d"   # Set the dataset name dynamically

# Define paths for the execution flag and GA result
flag_file <- paste0("ga_executed_flag_", dataset_name, ".RData")
result_file <- paste0("ga_result_", dataset_name, ".RData")

# Check if the result file exists
if (!file.exists(result_file)) {  
  message("No result file found. Running the Genetic Algorithm (GA).")
  
  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

  # Register start time
  start_time <- Sys.time()

  # Run the Genetic Algorithm
  ga_result <- ga(
    type = "binary",                  
    fitness = fitness_ga_svme,       
    nBits = num_features,             # Number of features
    popSize = 20,                     # Population size
    maxiter = 50,                     # Maximum iterations
    run = 10,                         # Convergence stopping criterion
    names = colnames(train_data[, -ncol(train_data)]),  # Feature names
    pcrossover = 0.8,                 # Probability of crossover
    pmutation = 0.1,                  # Probability of mutation
    maxFitness = -0.9,                # Stop when fitness reaches target
    parallel = TRUE,                  # Enable parallel processing
    monitor = gaMonitor,              # Track progress
    seed = 123                        # Reproducibility
  )

  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the GA model:', end_time - start_time))

  # Save the GA result and execution flag
  ga_executed <- TRUE
  save(ga_executed, ga_result, file = result_file)  # Save both flag and results
  
  message("GA execution completed and results saved.")
  
} else {
  # Load previously saved results
  message("Result file found. Loading saved GA results.")
  load(result_file)
  message(paste("GA already executed for dataset:", dataset_name))
}

```

## Visualization Plot

```{r}

par (cex = 0.8)
plot(ga_result, main = "GA Optimization Progress")
par(cex = 1)

```

## Analyze GA results

```{r}
summary(ga_result)
```

## Post-Selection

```{r}

# Get the best feature indices selected by GA
best_features <- which(ga_result@solution == 1)

# Remove indices beyond available columns
best_features <- best_features[best_features <= ncol(train_data)]

# Get selected feature names
selected_feature_names <- colnames(train_data)[best_features]

print(selected_feature_names)

```

```{r}
print(paste('The number of selected variables are', length(selected_feature_names)))
n_selected <- as.numeric(length(selected_feature_names))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, selected_feature_names, drop = FALSE]  
X_test_selected <- X_test[, selected_feature_names, drop = FALSE]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(X_train_selected, binary_class = y_train_selected)
test_merged <- cbind(X_test_selected, binary_class = y_test_selected)

dim(train_merged)
dim(test_merged)
table(train_merged$binary_class)
table(test_merged$binary_class)
```

## Create a Data frame with the Selected Features for SMOTE Data

```{r}
X_train_selected_smote <- X_train_smote[,selected_feature_names, drop = FALSE]
X_test_selected_smote <- X_test[, selected_feature_names, drop = FALSE] 

# Keep y_train and y_test as-is 
y_train_selected_smote <- y_train
y_test_selected_smote <- y_test

#Merge data_frame
train_merged_smote <- cbind(X_train_selected_smote, binary_class = smote_data$binary_class)
test_merged_smote <- cbind(X_test_selected_smote, binary_class = y_test_selected_smote)

dim(train_merged_smote)
dim(test_merged_smote)
table(train_merged_smote$binary_class)
table(test_merged_smote$binary_class)
```

## Check correlations by HeatMaps

```{r}
# Calculate the correlation matrix 

#Imbalanced Data

#Calculate correlations
corr <- cor(train_merged[, -(n_selected+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("blue", "white", "red"))(5), 
       title = "Correlation", cex = 0.8) 

```

```{r}
# SMOTE Data
par(cex.main = 0.6)
corr <- cor(train_merged_smote[, -(n_selected+1)])
# Create heat map
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("#440154", "#21908C", "gold"))(50),
        main = 'Correlation Heatmap',cex.main = 0.6)  

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("#440154", "#21908C", "gold"))(5), 
       title = "Correlation", 
       cex = 0.8) 
```

-   The correlation heat maps reveal that some variables remain highly correlated. To address this issue and mitigate multicollinearity, we will employ Principal Component Analysis (PCA)

## PreProcess by PCA, Centering and Scaling Data

```{r}

# Imbalanced Data
pca_Preprocess <- preProcess(train_merged, method = c("center", "scale", "pca"))
train_mergedPreprocessed <- predict(pca_Preprocess, train_merged) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,test_merged) # apply on test data

# SMOTE Data
pca_Preprocess_smote <- preProcess(train_merged_smote,method = c("center", "scale", "pca"))
train_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,train_merged_smote)
test_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,test_merged_smote)

# Check dimensions of the preprocessed data to ensure consistency

dim(train_mergedPreprocessed)
dim(test_mergedPreprocessed)

dim(train_merged_smote_Preprocessed)
dim(test_merged_smote_Preprocessed)
```

## Train Models

### Support Vector Machines (SVM)

- We will train SVM model using sparse variables selected by the svmE and GA algorithms, further reduced through Principal Component Analysis (PCA).


## Set Up Training Parameters

```{r}
# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# OneSE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```


```{r}
# Hyperparameter Tuning
# We will use tuneLength to automatically determine the optimal parameters Cost (C) and  RBF kernel γ parameter
```

## Train the Models

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_svm_msc_sg_1d<-train(as.factor(binary_class) ~.,
                         data = train_mergedPreprocessed,
                         method = 'svmRadial', tuneLength =  10,
                         trControl = control, 
                         metric = 'Accuracy') #unmodified data
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()
fit_svm_smote_msc_sg_1d <- train(as.factor(binary_class) ~., data = train_merged_smote_Preprocessed, 
                         method = 'svmRadial', tuneLength = 10,
                         trControl = control, 
                         metric = 'Accuracy') # SMOTE data

#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# Display model results for imbalanced data
print(fit_svm_msc_sg_1d)
```

```{r}
# Display results for data balanced by SMOTE
print(fit_svm_smote_msc_sg_1d)
```

## Save the models for future use

```{r}
saveRDS(fit_svm_msc_sg_1d, "fit_svm_msc_sg_1d.rds")
saveRDS(fit_svm_smote_msc_sg_1d, "fit_svm_smote_msc_sg_1d.rds")
```

## Plot the cross-validation results ----> Model selection

```{r}
# Imbalanced Data
fit_svm_msc_sg_1d
C_data<-fit_svm_msc_sg_1d$results
best_C = fit_svm_msc_sg_1d$bestTune$C
y_Acc = C_data$Accuracy[C_data$C == best_C]
plot(C_data$C,C_data$Accuracy, type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     xlab = "Cost", ylab = "Accuracy (%)",cex.lab = 0.9)
# Add line from x-axis to the specific point
segments(x0 = best_C, y0 = 0, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)
```

```{r}
# Balanced Data
fit_svm_smote_msc_sg_1d
C_data<-fit_svm_smote_msc_sg_1d$results
best_C = fit_svm_smote_msc_sg_1d$bestTune$C
y_Acc = C_data$Accuracy[C_data$C == best_C]
plot(C_data$C,C_data$Accuracy, type = "b",col = "blue",pch = 20,cex.axis = 0.8, 
     xlab = "Cost", ylab = "Accuracy (%)",cex.lab = 0.9,ylim = c(0,1))
# Add line from x-axis to the specific point
segments(x0 = best_C, y0 = 0, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)
```

## Display the cross-validation results

```{r}
print(paste('The optimal parametrers for training the SVM model are sigma value of', round(as.numeric(fit_svm_msc_sg_1d$bestTune$sigma),2), 'and Cost parameter of', fit_svm_msc_sg_1d$bestTune$C)) # unmodified data 
 
```


```{r}
print(paste('The optimal parametrers for training the SVM model are sigma value of', round(as.numeric(fit_svm_smote_msc_sg_1d$bestTune$sigma),2), 'and Cost parameter of', fit_svm_smote_msc_sg_1d$bestTune$C)) # SMOTE data
```

## Display CV results in Tables

```{r}
# Imbalanced data
kable(fit_svm_msc_sg_1d$results, caption = "SVM Cross-validation Results")
```

```{r}
# Select the best results row # unmodified data
selected_row <- fit_svm_msc_sg_1d$results %>% 
  filter(C==fit_svm_msc_sg_1d$bestTune$C & sigma == fit_svm_msc_sg_1d$bestTune$sigma)

kable(selected_row)
```

```{r}
# SMOTE Data
kable(fit_svm_smote_msc_sg_1d$results, caption = "SVM SMOTE Cross-validation Results")
```

```{r}
# Select the best results row 
selected_row <- fit_svm_smote_msc_sg_1d$results %>% 
  filter(C==fit_svm_smote_msc_sg_1d$bestTune$C & sigma == fit_svm_smote_msc_sg_1d$bestTune$sigma)

kable(selected_row)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_svm_msc_sg_1d$pred$pred,fit_svm_msc_sg_1d$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_svm_smote_msc_sg_1d$pred$pred,fit_svm_smote_msc_sg_1d$pred$obs) # SMOTE data
```

## Test the model performance on test data for unbalanced trained data

```{r}
# Predict on the test set 
set.seed(123)

test_svm_unmodified <- predict(fit_svm_smote_msc_sg_1d,newdata=test_mergedPreprocessed)

# Get the confusion matrix
test_svm_unmodified_CMtrx <- confusionMatrix(test_svm_unmodified,
                                             test_mergedPreprocessed$binary_class)
test_svm_unmodified_CMtrx

knitr::kable(test_svm_unmodified_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_svm_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm <- test_svm_unmodified_CMtrx$table
TP <- cf_table_svm[1,1]
TN <- cf_table_svm[2,2] 
FP <- cf_table_svm[2,1] 
FN <- cf_table_svm[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Predict on the test set 
set.seed(123)
test_svm_smote <- predict(fit_svm_smote_msc_sg_1d,
                          newdata=test_merged_smote_Preprocessed)

# Get the confusion matrix
test_svm_smote_CMtrx <- confusionMatrix(test_svm_smote,
                                             test_merged_smote_Preprocessed$binary_class)

test_svm_smote_CMtrx

knitr::kable(test_svm_smote_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_svm_smote_CMtrx$table)
```

```{r}
# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm_smote <- test_svm_smote_CMtrx$table
TP <- cf_table_svm_smote[1,1]
TN <- cf_table_svm_smote[2,2] 
FP <- cf_table_svm_smote[2,1] 
FN <- cf_table_svm_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```


## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(svm = fit_svm_msc_sg_1d, svm_smote = fit_svm_smote_msc_sg_1d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model performance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_svm_unmodified_CMtrx$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_smote <- as.data.frame(test_svm_smote_CMtrx$table)
plt_smote$Reference <- factor(plt_smote$Reference, levels=rev(levels(plt_smote$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: svm Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_smote, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: svm SMOTE Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(test_merged_smote_Preprocessed$binary_class,
                 as.numeric(test_svm_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("svm"), col = 'blue', lwd = 3)
```

### Misclassifications

```{r}
## Imbalanced
observed <- y_test
predicted <- test_svm_unmodified
perc <- hsi_msc_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_imbal <- imbal_class %>% filter(observed!=predicted)
misclas_imbal
```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# SMOTE Data
observed <- y_test
predicted <- test_svm_smote
perc <- hsi_msc_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

bal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_smote <- bal_class %>% filter(observed!=predicted)
misclas_smote
```

```{r}
table(misclas_smote$perc_adulter)
```
