---
title: "Ridge Regression with Boruta Feature Selection for Quantifying Coffee Adulteration: SNV+SG+1st Derivative Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "sky"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

Coffee adulteration, such as blending premium Arabica with lower-cost Robusta, undermines quality and authenticity, posing significant challenges for consumers, producers, and regulators. High-dimensional spectroscopic data provides valuable insights for detecting such fraud, but the large number of variables necessitates effective techniques to manage dimensionality and isolate the most relevant wavelengths.

This study combines **Boruta** feature selection with **Ridge Regression** to address these challenges. Boruta identifies and retains the most relevant spectral features by eliminating irrelevant and redundant variables, simplifying the dataset. Ridge Regression, applying L2 regularization, effectively handles any remaining multi-collinearity among the selected features while minimizing overfitting. This combined approach ensures robust modeling of high-dimensional spectral data, enhances prediction accuracy, and improves the detection of coffee adulteration.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(glmnet)
  library(viridis)
  library(Boruta)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: snv_sg_1d spectra

hsi_snv_sg_1d <- read.csv("coffee_instant_snv_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_snv_sg_1d)

# Check a few column names
col_check <- colnames(hsi_snv_sg_1d[,c(1:6)])
col_check

# our response variable (adulteration percentage) is column 4
# The spectral values start from column # 6 -------> 229

# Convert the adulteration percentage to numeric for compatibility with ML
hsi_snv_sg_1d$perc_adulter <- as.numeric(hsi_snv_sg_1d$perc_adulter)
```

## Get the X and y variables

```{r}
X <- hsi_snv_sg_1d[,c(6:229)]
y <- hsi_snv_sg_1d[,5]

class(y) #Ensure y (target variable) is a numeric for regression
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue', cex.main = 1.0, 
     cex.axis = 0.8,cex.lab = 1.0, freq= FALSE,
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()                                    
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_snv_sg_1d$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, y_train)
test_data <- data.frame(X_test, y_test)
```

## Run Boruta Algorithm for Optimum Wavelength Selection

```{r}

  # Set seed for reproducibility 
  set.seed(123)

  # Set the dataset name dynamically
  dataset_name <- "snv_sg_1d"  

  # Define paths for the execution flag and Boruta result
  flag_file <- paste0("boruta_executed_flag_", dataset_name, ".RData")
  result_file <- paste0("boruta_result_", dataset_name, ".RData")

  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

if (!file.exists(result_file)) {
  message(paste("Starting Boruta execution for dataset:", dataset_name))
  
  # Start the timer
  start_time <- Sys.time()

  # Run Boruta on the train data set
  boruta_result <- Boruta(formula = y_train ~ .,  
                        data = train_data, 
                        doTrace = 2,
                        maxRuns = 100)                   
 
  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the Boruta model:', end_time - start_time))  
  
  # Save Boruta results and execution flag
  boruta_executed <- TRUE
  save(boruta_executed, file = flag_file)
  save(boruta_result, file = result_file)
 

  message(paste("Boruta execution completed for dataset:", dataset_name, "and results saved."))

} else {
  # Load previously saved results
  message(paste("Result file found. Loading saved Boruta results for dataset:", dataset_name))
  load(result_file)
  load(flag_file)
  message(paste("Boruta already executed for dataset:", dataset_name))
}

```


```{r}
# Print Boruta summary
print(boruta_result)
```

```{r}
# Get selected features
final_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
final_features
```


```{r}
# Plot Boruta results
plot(boruta_result, xlab = "", las = 2, 
     cex.main = 0.8, cex.axis = 0.6, 
     main = paste("Boruta Feature Selection"))
```

```{r}
print(paste('The number of selected variables are', length(final_features)))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, final_features]  
X_test_selected <- X_test[, final_features]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(y_train = y_train_selected, X_train_selected)
test_merged <- cbind(y_test = y_test_selected, X_test_selected)

dim(train_merged)
dim(test_merged)
```


## Check correlations by HeatMaps

```{r}

n_features <- length(final_features)

#Calculate correlations
corr <- cor(train_merged[, -(n_features+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = viridis(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = viridis(5), 
       title = "Correlation", cex = 0.8) 
```

## Model Training Set Up

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        savePredictions = "final", 
                        summaryFunction = defaultSummary,
                        selectionFunction = 'best', allowParallel = TRUE)

# We used oneSE for classification. For regression, we will use 'best': a model that minimizes RMSECV
```

```{r}
#Set the metric as accuracy 
metric<-"RMSE"
```

```{r}
# Hyperparameter set
# alpha: Set to 0 for Ridge (pure L2 regularization).
# lambda: Regularization strength (penalty parameter). A sequence of values is tested to identify the best lambda.

grid <- expand.grid(alpha = 0, lambda = 10^seq(-3, 3, length = 100)) 
```

# Train Ridge Regression Model

```{r}
set.seed(123)
#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_ridge_snv_sg_1d<-train(y = train_merged[,1], 
                         x = train_merged[,-1],
                         method = 'glmnet', tuneGrid = grid,
                         trControl = control, preProcess = c("center","scale"),
                         metric = metric) 
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
# Display regression results
print(fit_ridge_snv_sg_1d)
```
#### Best hyperparameters
```{r}
# Best lambda value
best_lambda <- fit_ridge_snv_sg_1d$bestTune$lambda

cat("Best Lambda:", best_lambda, "\n")
```
## Save the models for future use

```{r}
saveRDS(fit_ridge_snv_sg_1d, "fit_ridge_snv_sg_1d.rds")
```

## Plot the cross-validation results —–> Model Selection 

```{r}
# Extract results and best lambda
ridge_data <- fit_ridge_snv_sg_1d$results
best_lambda <- fit_ridge_snv_sg_1d$bestTune$lambda
y_RMSE <- ridge_data$RMSE[ridge_data$lambda == best_lambda]

# Plot RMSE against lambda
plot(
  ridge_data$lambda, ridge_data$RMSE, type = "b", col = "black", pch = 8, cex.axis = 0.8,
  xlab = "Lambda", ylab = "RMSECV(%)", cex.lab = 0.9,
  main = "Ridge Optimal Model Selection", cex.main = 0.9, cex.lab = 0.8
)

# Add line from x-axis to the specific point (optimal lambda)
segments(x0 = best_lambda, y0 = 0, x1 = best_lambda, y1 = y_RMSE, col = "red", lty = 2)

# Add line from y-axis to the specific point (corresponding RMSE)
segments(x0 = 0, y0 = y_RMSE, x1 = best_lambda, y1 = y_RMSE, col = "red", lty = 2)

# Add annotation for the optimal point
text(best_lambda, y_RMSE, labels = paste0("(", round(best_lambda, 4), ", ", round(y_RMSE, 2), ")"), 
     pos = 3, col = "blue", cex = 0.8)

```

#### Coefficients vs. Lambda Path

```{r}
# Extract the final glmnet model 
fit_glmnet <- fit_ridge_snv_sg_1d$finalModel

# Plot coefficients vs. lambda
par(cex.main = 0.8)
plot(
  fit_glmnet, 
  xvar = "lambda",      
  label = FALSE,         
  cex.axis = 0.8,       
  cex.lab = 0.9,        
  cex.main = 0.3)
title(main = "Coefficient Path for Ridge Regression")


#As lambda increases, coefficients shrink toward zero, highlighting the feature selection process
```

## Display CV Results in Tables

```{r}
kable(fit_ridge_snv_sg_1d$results)
```


```{r}
# Select row for the optimal model
opt_row <- fit_ridge_snv_sg_1d$results %>% 
  filter(lambda==fit_ridge_snv_sg_1d$bestTune$lambda)

kable(opt_row)
```


## Test the model performance on test data 

```{r}
#Predict on the test set 
observed <- test_data$y_test
predictions <- predict(fit_ridge_snv_sg_1d,newdata=test_data)

# Generate predictions
results <- postResample(predictions, observed)

print(results)

RMSEP = results[[1]]
R2 = results[[2]]
MAE = results[[3]]

cat('The RMSEP for the model on external validation is:', RMSEP,'\n')
cat('The R Squared value for the model is:', R2)

# Calculate Residual Predictive Deviation to check model performance

RPD = sd(observed)/RMSEP

print(paste('The RPD values for the model is', round(RPD,2)))
```

## Visualizations

```{r}

# predicted vs observed line and scatter
# Create a data frame
results_df <- data.frame(observed=observed, predictions=predictions)

results_df %>%
  ggplot(aes(x = observed, y = predictions)) +
  geom_point(color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", 
              linetype = "dashed", linewidth = 1) +
  labs( x = "Measured Adulteration (%)",
    y = "Predicted Adulteration (%)") +
  annotate("text",
    x = min(observed), y = max(predictions),
    label = paste("RMSEP:", round(RMSEP, 2),
      "\nR-squared:", round(R2, 3),
      "\nRPD:", round(RPD, 2)),
    hjust = 0, vjust = 1, size = 3, color = "black") +
  theme(
    axis.title = element_text(color = "black", size = 10),
    axis.text = element_text(color = "black"))+
  theme_gray()

```

```{r}
# Check residuals

residuals <- observed - predictions

# Plot Residuals
plot(predictions, residuals,
     xlab = "Predicted Values",
     ylab = "Residuals",
     main = "Residuals vs Predicted",
     pch = 16, col = "blue")
abline(h = 0, col = "red", lwd = 2)  
```


```{r}
hist(residuals, 
     main = "Histogram of Residuals",
     xlab = "Residuals", cex.main = 0.8,
     col = "lightblue", border = "black"); box()

```

