---
title: "Elastic Net Regression for Quantifying Coffee Adulteration: Raw Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

Coffee adulteration, such as blending premium Arabica with lower-cost Robusta, undermines quality and authenticity, posing significant challenges for consumers, producers, and regulators. High-dimensional spectroscopic data provides valuable insights for detecting such fraud, but the large number of variables necessitates effective techniques to manage dimensionality and isolate the most relevant wavelengths.

This study employs Elastic Net regression, a robust technique that combines the strengths of elastic_net (L1 regularization) and Ridge (L2 regularization). Elastic Net balances feature selection and coefficient shrinkage, effectively addressing multicollinearity while retaining critical spectral features. This approach enables efficient modeling with high-dimensional data by reducing redundancy and improving prediction accuracy.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(glmnet)
  library(viridis)
  library(Boruta)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: raw spectra

hsi_raw <- read.csv("coffee_data_pycleaned.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_raw)

# Check a few column names
col_check <- colnames(hsi_raw[,c(1:6)])
col_check

# our response variable (adulteration percentage) is column 4
# The spectral values start from column # 6 -------> 229

# Convert the adulteration percentage to numeric for compatibility with ML
hsi_raw$adult_percent <- as.numeric(hsi_raw$adult_percent)
```

## Get the X and y variables

```{r}
X <- hsi_raw[,c(6:229)]
y <- hsi_raw[,4]

class(y) #Ensure y (target variable) is a numeric for regression
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue', cex.main = 1.0, cex.axis = 0.8,cex.lab = 1.0,freq = FALSE,
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()                                    
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_raw$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(y_train, X_train)
test_data <- data.frame(y_test, X_test)
```

## Train Models

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        savePredictions = "final", 
                        summaryFunction = defaultSummary,
                        selectionFunction = 'best', allowParallel = TRUE)

# We used oneSE for classification. For regression, we will use 'best': a model that minimizes RMSECV
```

```{r}
#Set the metric as accuracy 
metric<-"RMSE"
```

```{r}
# Hyperparameter set Elastic Net Regression
# alpha: Set for L1(elastic_net) and L2(Ridge) Regression
# lambda: Regularization strength (penalty parameter). A sequence of values is tested to identify the best lambda.

grid <- expand.grid( alpha = seq(0, 1, length = 10),  # Explore between Ridge (alpha = 0) and elastic_net (alpha = 1)
    lambda = seq(0.001, 0.1, length = 10))
```

# Train Elastic Net

```{r}
set.seed(123)
#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_elastic_net_raw<-train(y = train_data[,1], 
                         x = train_data[,-1],
                         method = 'glmnet', tuneGrid = grid,
                         trControl = control, preProcess = c("center","scale"),
                         metric = metric) 
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
# Display regression results
print(fit_elastic_net_raw)
```
#### Best hyperparameters
```{r}
# Best lambda value
best_lambda <- fit_elastic_net_raw$bestTune$lambda

cat("Best Lambda:", best_lambda, "\n")
```

#### Extract Selected Features

```{r}
# Extract coefficients from the best model
coefficients <- coef(fit_elastic_net_raw$finalModel, s = best_lambda)

# Convert to a readable format
elastic_net_coefficients <- as.data.frame(as.matrix(coefficients))
colnames(elastic_net_coefficients) <- "Coefficient"
elastic_net_coefficients$Variable <- rownames(elastic_net_coefficients)

# Filter non-zero coefficients (selected features)
selected_features <- elastic_net_coefficients[elastic_net_coefficients$Coefficient != 0, ]
print(selected_features)

# Sort the selected features by actual values in descending order
selected_features_sorted_actual <- selected_features[order(selected_features$Coefficient, decreasing = TRUE), ]

# Print the sorted features
print(selected_features_sorted_actual)

```

```{r}
print(paste('The number of feautures selected by elastic_net are', nrow(selected_features_sorted_actual)))
```
## Save the models for future use

```{r}
saveRDS(fit_elastic_net_raw, "fit_elastic_net_raw.rds")
```

#### Coefficients vs. Lambda Path

```{r warning=FALSE}
# Extract the final glmnet model 
fit_glmnet <- fit_elastic_net_raw$finalModel

# Plot coefficients vs. lambda
par(cex.main = 0.8)
plot(
  fit_glmnet, 
  xvar = "lambda",      
  label = FALSE,         
  cex.axis = 0.8,       
  cex.lab = 0.9,        
  cex.main = 0.3)
title(main = "Coefficient Path for elastic_net Regression")


#As lambda increases, coefficients shrink toward zero, highlighting the feature selection process
```

## Display CV Results in Tables

```{r}
kable(fit_elastic_net_raw$results)
```


```{r}
# Select row for the optimal model
opt_row <- fit_elastic_net_raw$results %>% 
  filter(lambda==fit_elastic_net_raw$bestTune$lambda)

kable(opt_row)
```


## Test the model performance on test data 

```{r}
#Predict on the test set 
observed <- test_data$y_test
predictions <- predict(fit_elastic_net_raw,newdata=test_data)

# Generate predictions
results <- postResample(predictions, observed)

print(results)

RMSEP = results[[1]]
R2 = results[[2]]
MAE = results[[3]]

cat('The RMSEP for the model on external validation is:', RMSEP,'\n')
cat('The R Squared value for the model is:', R2)

# Calculate Residual Predictive Deviation to check model performance

RPD = sd(observed)/RMSEP

print(paste('The RPD values for the model is', round(RPD,2)))
```

## Visualizations

```{r}

# predicted vs observed line and scatter
# Create a data frame
results_df <- data.frame(observed=observed, predictions=predictions)

results_df %>%
  ggplot(aes(x = observed, y = predictions)) +
  geom_point(color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", 
              linetype = "dashed", linewidth = 1) +
  labs( x = "Measured Adulteration (%)",
    y = "Predicted Adulteration (%)") +
  annotate("text",
    x = min(observed), y = max(predictions),
    label = paste("RMSEP:", round(RMSEP, 2),
      "\nR-squared:", round(R2, 3),
      "\nRPD:", round(RPD, 2)),
    hjust = 0, vjust = 1, size = 3, color = "black") +
  theme(
    axis.title = element_text(color = "black", size = 10),
    axis.text = element_text(color = "black"))+
  theme_gray()

```

```{r}
# Check residuals

residuals <- observed - predictions

# Plot Residuals
plot(predictions, residuals,
     xlab = "Predicted Values",
     ylab = "Residuals",
     main = "Residuals vs Predicted",
     pch = 16, col = "blue")
abline(h = 0, col = "red", lwd = 2)  
```


```{r}
hist(residuals, 
     main = "Histogram of Residuals",
     xlab = "Residuals", cex.main = 0.8,
     col = "lightblue", border = "black"); box()

```


## Conclusion

- Elastic_net demonstrate good performance  in predicting adulteration in ground coffee. The model achieves an RPD value of 2.8. The model can make reliable predictions and is suitable for quality control.