---
title: "Feature Optimization for Coffee Adulteration Detection Using Hyperspectral Imaging and Ensemble Machine Learning: Integrating RFE and GA with MSC+SG+2nd Derivative Preprocessing"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: true
---

## Introduction

This project explores the effectiveness of ensemble learning methods in distinguishing authentic coffee from adulterated samples using spectral data derived from hyperspectral imaging (HSI). By incorporating feature selection techniques such as Recursive Feature Elimination (RFE) and Genetic Algorithms (GA), the study identifies the most relevant spectral features, significantly reducing dimensionality while maintaining or enhancing model accuracy. These feature selection methods help mitigate the curse of dimensionality, ensuring the ensemble models focus on the most informative wavelengths.

Ensemble learning combines predictions from multiple base models, including Random Forest (RF), Support Vector Machines (SVM), Artificial Neural Networks (ANN), Linear Discriminant Analysis (LDA), and k-Nearest Neighbors (kNN), to improve robustness and classification accuracy. By leveraging stacking and voting approaches, the study capitalizes on the strengths of diverse learning algorithms, ensuring better generalization to varying adulteration levels.

Preprocessing techniques such as Multiplicative Scatter Correction (MSC), Savitzky-Golay (SG) smoothing, and second derivative transformations are applied to enhance spectral quality and remove noise. The ensemble's performance is evaluated using robust cross-validation techniques, including 10-fold repeated cross-validation, to ensure reliability and prevent overfitting.

The study focuses on classifying pure and adulterated coffee across varying adulteration levels (e.g., 1%, 5%, 10%, and higher concentrations). Additionally, the impact of feature selection and ensemble learning on computational efficiency, sensitivity, and specificity is critically analyzed, offering a comprehensive framework for using ensemble models in food authenticity and fraud detection studies.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis) 
  library(NeuralNetTools)
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(caretEnsemble)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(ggthemes)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: msc_sg_2d spectra

hsi_msc_sg_2d <- read.csv("coffee_ground_msc_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_2d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6 -------> 229

# Convert the class to a factor
hsi_msc_sg_2d$binary_class <- as.factor(hsi_msc_sg_2d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_msc_sg_2d[,c(6:229)]
y <- hsi_msc_sg_2d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'steelblue1', freq = FALSE,
     main = 'Histogram of Spectral Intensity Values', cex.main = 0.8,
     xlab = 'Intensity'); box()
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample <- hsi_msc_sg_2d$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <- y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("slateblue", "sienna")) +  # Specify your custom colors here
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +  # Ensure legend names are retained
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 10))

```

## Balance the data using SMOTE

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the X and y for the over-sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("darkcyan", "slateblue")) +  # Blue and Orange
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,size = 10))

```

## Define the Fitness Function for GA

- This function evaluates subsets of features selected by GA using RFE with a Random Forest (RF) model.

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}

# Fitness Function
fitness_ga_rfe <- function(subset) {
  # Identify selected features based on the binary subset
  selected_features <- which(subset == 1)
  
  # Penalize empty subsets to avoid errors
  if (length(selected_features) == 0) return(Inf)
  
  # Subset the training feature matrix to selected features
  X_selected <- X_train[, selected_features]
  
  # Define control for rfe
  control <- rfeControl(functions = rfFuncs,  # Use Random Forest for rfe
                        method = "cv",       # Cross-validation
                        number = 10)         # 10-fold CV
  
  # Apply rfe to the selected subset of features
  rfe_result <- rfe(X_selected, y_train, rfeControl = control)
  
  # Return the negative accuracy as the fitness score (to maximize accuracy)
  return(-rfe_result$results$Accuracy[which.max(rfe_result$results$Accuracy)])
}
```

## Run the Genetic Algorithm

```{r}

# Set seed for reproducibility 
set.seed(123)

num_features <- ncol(X_train) # Number of features
dataset_name <- "msc_sg_2d"   # Set the dataset name dynamically

# Define paths for the execution flag and GA result
flag_file <- paste0("ga_executed_flag_", dataset_name, ".RData")
result_file <- paste0("ga_result_", dataset_name, ".RData")

# Check if the result file exists
if (!file.exists(result_file)) {  
  message("No result file found. Running the Genetic Algorithm (GA).")
  
  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

  # Register start time
  start_time <- Sys.time()

  # Run the Genetic Algorithm
  ga_result <- ga(
    type = "binary",                  
    fitness = fitness_ga_rfe,       
    nBits = num_features,             # Number of features
    popSize = 20,                     # Population size
    maxiter = 50,                     # Maximum iterations
    run = 10,                         # Convergence stopping criterion
    names = colnames(train_data[, -ncol(train_data)]),  # Feature names
    pcrossover = 0.8,                 # Probability of crossover
    pmutation = 0.1,                  # Probability of mutation
    maxFitness = -0.9,                # Stop when fitness reaches target
    parallel = TRUE,                  # Enable parallel processing
    monitor = gaMonitor,              # Track progress
    seed = 123                        # Reproducibility
  )

  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the GA model:', end_time - start_time))

  # Save the GA result and execution flag
  ga_executed <- TRUE
  save(ga_executed, ga_result, file = result_file)  # Save both flag and results
  
  message("GA execution completed and results saved.")
  
} else {
  # Load previously saved results
  message("Result file found. Loading saved GA results.")
  load(result_file)
  message(paste("GA already executed for dataset:", dataset_name))
}

```

## Visualization Plot

```{r}

par (cex = 0.8)
plot(ga_result, main = "GA Optimization Progress")
par(cex = 1)

```

## Analyze GA results

```{r}
summary(ga_result)
```

## Post-Selection

```{r}
# Get the best feature subset
best_features <- which(ga_result@solution == 1)
selected_feature_names <- colnames(train_data)[best_features]
print(selected_feature_names)
```


```{r}
print(paste('The number of selected variables are', length(selected_feature_names)))
n_selected <- as.numeric(length(selected_feature_names))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, selected_feature_names, drop = FALSE]  
X_test_selected <- X_test[, selected_feature_names, drop = FALSE]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(X_train_selected, binary_class = y_train_selected)
test_merged <- cbind(X_test_selected, binary_class = y_test_selected)

dim(train_merged)
dim(test_merged)
table(train_merged$binary_class)
table(test_merged$binary_class)
```

## Create a Data frame with the Selected Features for SMOTE Data

```{r}
X_train_selected_smote <- X_train_smote[,selected_feature_names, drop = FALSE]
X_test_selected_smote <- X_test[, selected_feature_names, drop = FALSE] 

# Keep y_train and y_test as-is 
y_train_selected_smote <- y_train
y_test_selected_smote <- y_test

#Merge data_frame
train_merged_smote <- cbind(X_train_selected_smote, binary_class = smote_data$binary_class)
test_merged_smote <- cbind(X_test_selected_smote, binary_class = y_test_selected_smote)

dim(train_merged_smote)
dim(test_merged_smote)
table(train_merged_smote$binary_class)
table(test_merged_smote$binary_class)
```

## Check correlations by HeatMaps

```{r}
# Calculate the correlation matrix 

#Imbalanced Data

#Calculate correlations
corr <- cor(train_merged[, -(n_selected+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("blue", "white", "red"))(5), 
       title = "Correlation", cex = 0.8) 

```

```{r}
# SMOTE Data
par(cex.main = 0.6)
corr <- cor(train_merged_smote[, -(n_selected+1)])
# Create heat map
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("#440154", "#21908C", "gold"))(50),
        main = 'Correlation Heatmap',cex.main = 0.6)  

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("#440154", "#21908C", "gold"))(5), 
       title = "Correlation", 
       cex = 0.8) 
```

-   The correlation heat maps reveal that some variables remain highly correlated. To address this issue and mitigate multicollinearity, we will employ Principal Component Analysis (PCA). We will then proceed to train our models with features generated by PCA. This aims to speed up processing time.

## PreProcess by PCA, Centering and Scaling Data

```{r}

# Imbalanced Data
pca_Preprocess <- preProcess(train_merged, method = c("center", "scale", "pca"))
train_mergedPreprocessed <- predict(pca_Preprocess, train_merged) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,test_merged) # apply on test data

# SMOTE Data
pca_Preprocess_smote <- preProcess(train_merged_smote,method = c("center", "scale", "pca"))
train_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,train_merged_smote)
test_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,test_merged_smote)

# Check dimensions of the preprocessed data to ensure consistency

dim(train_mergedPreprocessed)
dim(test_mergedPreprocessed)

dim(train_merged_smote_Preprocessed)
dim(test_merged_smote_Preprocessed)
```

## Set Up Training Parameters

```{r}

# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# One SE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# We will train all the models: kNN, NNET, SVM, LDA and RF as base learners. Then the predictions from this model will be trained by RF as a  meta-learner.

# Hyperparameter Tuning: We will train the model with their own set of hyper parameter settings.

# LDA model: No hyperparameter

# kNN
grid_knn <- expand.grid(.k = seq(3,30, by = 2))

# SVM. We will use a tuneLength of 10. This will automatically select optimal Cost (C) and sigma parameter for the RBF kernel 

# RF: We will also use a tuneLength of 10. This selects the optimal mtry/number of splits

# NNET model. Selects the optimal number of neurons and decay value  (regularization parameter)
grid_nnet <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

## Train the Models

```{r cache=TRUE}

# Unmodified data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_msc_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'lda',trControl = control, metric = 'Accuracy')

# 2. Train the kNN model
fit_knn_model_msc_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'knn',trControl = control, 
                       tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_msc_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'svmRadial',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_msc_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'rf',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_msc_sg_2d<-train(binary_class ~., data = train_mergedPreprocessed, 
                      method = 'nnet', tuneGrid = grid_nnet,
                      trControl = control, metric = 'Accuracy',
                      trace = FALSE, MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# After training the models, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list = list(lda = fit_lda_model_msc_sg_2d, 
                   knn = fit_knn_model_msc_sg_2d,
                   svm = fit_svm_model_msc_sg_2d,
                   rf = fit_rf_model_msc_sg_2d,
                   nnet = fit_nnet_msc_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_msc_sg_2d <- caretStack(models_list, method = 'rf', metric = metric,
                                       trControl = control,tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```
```{r}
# Train the models on SMOTE balanced data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_smote_msc_sg_2d <- train(binary_class ~.,
                                       data = train_merged_smote_Preprocessed,
                                       method = 'lda',trControl = control, 
                                       metric = 'Accuracy')

# 2. Train the kNN model
fit_knn_model_smote_msc_sg_2d <- train(binary_class ~., 
                                 data = train_merged_smote_Preprocessed,
                                 method = 'knn',trControl = control, 
                                 tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_smote_msc_sg_2d <- train(binary_class ~., 
                                       data = train_merged_smote_Preprocessed,
                                       method = 'svmRadial',trControl = control,
                                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_smote_msc_sg_2d <- train(binary_class ~., 
                                data = train_merged_smote_Preprocessed,
                                method = 'rf',trControl = control,
                                tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_smote_msc_sg_2d<-train(binary_class ~., 
                          data = train_merged_smote_Preprocessed,
                          method = 'nnet', tuneGrid = grid_nnet,
                          trControl = control, metric = 'Accuracy',
                          trace = FALSE, MaxNWts = 1000, maxit = 300) 
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```



```{r}
# Similarly, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list_smote = list(lda = fit_lda_model_smote_msc_sg_2d, 
                         knn = fit_knn_model_smote_msc_sg_2d,
                         svm = fit_svm_model_smote_msc_sg_2d,
                         rf = fit_rf_model_smote_msc_sg_2d,
                         nnet = fit_nnet_smote_msc_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_smote_msc_sg_2d <- caretStack(models_list_smote, 
                                             method = 'rf', 
                                             metric = metric,
                                             trControl = control,
                                             tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
print(fit_stack_ensemble_msc_sg_2d) # Model on imbalanced data
```

```{r}
print(fit_stack_ensemble_smote_msc_sg_2d) # SMOTE model
```


## Save the models for future use

```{r}
saveRDS(fit_stack_ensemble_msc_sg_2d, 
        "fit_stack_ensemble_msc_sg_2d.rds") # for unmodified data

saveRDS(fit_stack_ensemble_smote_msc_sg_2d, 
        "fit_stack_ensemble_smote_msc_sg_2d.rds") # SMOTE data
```

## Display the optimal parameters for the ensemble meta learner

```{r}
# Used 500 trees as a default
print(paste('The optimal parametrers for training the meta-learner rf model (on unmodified data) are mtry value of', 
            fit_stack_ensemble_msc_sg_2d$ens_model$bestTune)) # unmodified data 

print(paste('The optimal parametrers for training the meta-learner rf model (on data balanced by SMOTE) are  mtry value of', fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune)) # unmodified data  # SMOTE data 
```
## Display the cross-validation results in Tables

```{r warning=FALSE}
# Imbalanced data
kable(fit_stack_ensemble_msc_sg_2d$ens_model$results, 
      caption = "Cross-Validation Results on Imbalanced Data") # Imbalanced Data
```


```{r warning=FALSE}
# SMOTE data
kable(fit_stack_ensemble_smote_msc_sg_2d$ens_model$results,
      caption = "Cross-Validation Results on SMOTE Data") # Balanced Data
```

```{r}
# Select the best results row for unmodified data
selected_row <- fit_stack_ensemble_msc_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_msc_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)
```

```{r}
# Select the best results row for SMOTE Data
selected_row <- fit_stack_ensemble_smote_msc_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)

```

```{r}
# Select the best results row 
# Select the best results row for unmodfied data

selected_row <- fit_stack_ensemble_msc_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_msc_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_stack_ensemble_msc_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_msc_sg_2d$ens_model$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_stack_ensemble_smote_msc_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_smote_msc_sg_2d$ens_model$pred$obs) # SMOTE data
```

Visualize Hyperparameter Tuning for Optimal Model Selection

```{r}

# Imbalanced Data

mtry_data <- fit_stack_ensemble_msc_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_msc_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

```{r}

# SMOTE Data

mtry_data <- fit_stack_ensemble_smote_msc_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_smote_msc_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

## Visualize the Out of Bag Errors (OOB)

```{r}

# Unmodified data

# Plot the OOB error from the ensemble model
plot(fit_stack_ensemble_msc_sg_2d$ens_model$finalModel, 
     main = 'OOB Error Rate', 
     cex.main = 0.8,
     cex.axis = 0.8,
     col = c('blue', 'red', 'black'),
     lwd = 1.5)

# Add legend
legend("topright",
       legend = c("OOB Error", "Adulterated", "Arabica"),
       col = c('blue', 'red', 'black'),
       lty = c(1,2,2),
       lwd = 1.5,
       cex = 0.8)
```

```{r}
# SMOTE Data 

# Plot the OOB error from the ensemble model
plot(fit_stack_ensemble_smote_msc_sg_2d$ens_model$finalModel, 
     main = 'OOB Error Rate', 
     cex.main = 0.8,
     cex.axis = 0.8,
     col = c('blue', 'red', 'black'),
     lwd = 1.5)

# Add legend
legend("topright",
       legend = c("OOB Error", "Adulterated", "Arabica"),
       col = c('blue', 'red', 'black'),
       lty = c(1,2,2),
       lwd = 1.5,
       cex = 0.8)
```

## Test the model peformance on test data for unbalanced trained data

```{r}
# Make predictions with the ensemble model: Imbalanced Data

test_ensemble_unmodified <- predict(fit_stack_ensemble_msc_sg_2d,
                                newdata=test_mergedPreprocessed)

# Apply threshold to probabilities
class_predictions_unmodified <- ifelse(test_ensemble_unmodified[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_unmodified <- factor(class_predictions_unmodified, 
                                       levels = c("adulterated", "pure_arabica"))

reference_unmodified <- test_mergedPreprocessed$binary_class

# Get the confusion matrix

test_ensemble_unmodified_cfmatrix <- confusionMatrix(class_predictions_unmodified,
                                          reference_unmodified)
test_ensemble_unmodified_cfmatrix

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_unmodified_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Make predictions with the ensemble model: SMOTE Data

test_ensemble_smote <- predict(fit_stack_ensemble_smote_msc_sg_2d,
                                newdata=test_merged_smote_Preprocessed)

# Apply threshold to probabilities
class_predictions_smote <- ifelse(test_ensemble_smote[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_smote <- factor(class_predictions_smote, 
                                       levels = c("adulterated", "pure_arabica"))

reference_smote <- test_merged_smote_Preprocessed$binary_class

# Get the confusion matrix

test_ensemble_smote_cfmatrix <- confusionMatrix(class_predictions_smote,
                                          reference_smote)
test_ensemble_smote_cfmatrix

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_smote_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(ensemble_1 = fit_stack_ensemble_msc_sg_2d$ens_model, 
                          ensemble_smote = fit_stack_ensemble_smote_msc_sg_2d$ens_model))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert re-sampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results, col = 'red')

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
p1<-ggplot(data = as.data.frame(test_ensemble_unmodified_cfmatrix$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "darkgoldenrod", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix Unbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(data = as.data.frame(test_ensemble_smote_cfmatrix$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "olivedrab", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r warning=FALSE}

# Generate the ROC curve
roc_curve <- roc(test_merged_smote_Preprocessed$binary_class,
                 as.numeric(class_predictions_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')

# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("Ensemble"), col = 'blue', lwd = 3)

```


## Conclusion

- The ensemble stacked models, incorporating variable reduction through Genetic Algorithm (GA) and Recursive Feature Elimination (RFE), demonstrate exceptional performance in classifying authentic and adulterated coffee samples. The technique achieves a Matthews Correlation Coefficient (MCC) value greater than 0.9, underscoring its robustness and reliability in distinguishing between pure Arabica and adulterated coffee with high accuracy and balanced predictions.
