---
title: "Partial Least Squares Regression (PLSR) for Quantifying Coffee Adulteration: Spectral Preprocessing Using SNV, Savitzky-Golay, and First Derivative"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

Coffee adulteration, such as blending premium Arabica with lower-cost Robusta, undermines quality and authenticity, posing significant challenges for consumers, producers, and regulators. High-dimensional spectroscopic data, encompassing the full spectrum, provides valuable insights for detecting such fraud. However, the large number of variables requires advanced techniques to effectively model the data without overfitting.

This study employs Partial Least Squares regression (PLS) to analyze the full spectral data set and predict the levels of Arabica coffee adulteration with Robusta, ranging from 1% to 80%. By leveraging the entire spectral range, the approach aims to maximize predictive accuracy while addressing the complexities of high-dimensional data.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(viridis)
  library(Boruta)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: snv_sg_1d spectra

hsi_snv_sg_1d <- read.csv("coffee_ground_snv_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_snv_sg_1d)

# Check a few column names
col_check <- colnames(hsi_snv_sg_1d[,c(1:6)])
col_check

# our response variable (adulteration percentage) is column 4
# The spectral values start from column # 6 -------> 229

# Convert the adulteration percentage to numeric for compatibility with ML
hsi_snv_sg_1d$adult_percent <- as.numeric(hsi_snv_sg_1d$adult_percent)
```

## Get the X and y variables

```{r}
X <- hsi_snv_sg_1d[,c(6:229)]
y <- hsi_snv_sg_1d[,4]

class(y) #Ensure y (target variable) is a numeric for regression
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue', cex.main = 1.0, cex.axis = 0.8,cex.lab = 1.0,
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()                                    
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_snv_sg_1d$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(y_train, X_train)
test_data <- data.frame(y_test, X_test)
```




## Check correlations by HeatMaps

```{r}

#Calculate correlations
corr <- cor(train_data[,-1])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = viridis(50),
        main = "Correlation Heatmap")

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = viridis(5), 
       title = "Correlation", cex = 0.8) 
```

## Train Models

-   We will train models based all spectral features

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        savePredictions = "final", 
                        summaryFunction = defaultSummary,
                        selectionFunction = 'best', allowParallel = TRUE)

# We used oneSE for classification. For regression, we will use 'best': a model that minimizes RMSECV
```

```{r}
#Set the metric as accuracy 
metric<-"RMSE"
```

```{r}
# Hyperparameter Tuning
# We will use tune Length to automatically determine the optimal number of PLS components
```

```{r}
set.seed(123)
#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_pls_snv_sg_1d<-train(y = train_data[,1], 
                         x = train_data[,-1],
                         method = 'pls', tuneLength = 50,
                         trControl = control, 
                         metric = metric) 
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
# Display regression results
print(fit_pls_snv_sg_1d)
```

## Save the models for future use

```{r}
saveRDS(fit_pls_snv_sg_1d, "fit_pls_snv_sg_1d.rds")
```

## Plot the cross-validation results —–> Model Selection 

```{r}
# Imbalanced Data
pls_data <- fit_pls_snv_sg_1d$results
best_Lvs = fit_pls_snv_sg_1d$bestTune$ncomp
y_RMSE = pls_data$RMSE[pls_data$ncomp == best_Lvs]
plot(pls_data$ncomp,pls_data$RMSE, type = "b",col = "black",pch = 8,cex.axis = 0.8, 
     xlab = "Latent Variables", ylab = "RMSECV(%)",cex.lab = 0.9,
     main = 'PLS Optimal Model Selection',
     cex.main = 0.9,
     cex.lab = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_Lvs, y0 = 0, x1 = best_Lvs, y1 = y_RMSE, col = "red", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_RMSE, x1 = best_Lvs, y1 = y_RMSE, col = "red", lty = 2)
```

## Display the cross-validation results

```{r}
print(paste('The optimal parameters for training the pls model is', 
            fit_pls_snv_sg_1d$bestTune$ncomp, 'components'))   
```

## Display CV Results in Tables

```{r}
kable(fit_pls_snv_sg_1d$results)
```


```{r}
# Select row for the optimal model
opt_row <- fit_pls_snv_sg_1d$results %>% 
  filter(ncomp==fit_pls_snv_sg_1d$bestTune$ncomp)

kable(opt_row)
```


## Test the model performance on test data 

```{r}
#Predict on the test set 
observed <- test_data$y_test
predictions <- predict(fit_pls_snv_sg_1d,newdata=test_data)

# Generate predictions
results <- postResample(predictions, observed)

print(results)

RMSEP = results[[1]]
R2 = results[[2]]
MAE = results[[3]]

cat('The RMSEP for the model on external validation is:', RMSEP,'\n')
cat('The R Squared value for the model is:', R2,'\n')

# Calculate Residual Predictive Deviation to check model performance

RPD = sd(observed)/RMSEP

print(paste('The RPD values for the model is', round(RPD,2)))
```

## Visualizations

```{r}

# predicted vs observed line and scatter
# Create a data frame
results_df <- data.frame(observed=observed, predictions=predictions)

results_df %>%
  ggplot(aes(x = observed, y = predictions)) +
  geom_point(color = "blue") +
  geom_abline(slope = 1, intercept = 0, color = "red", 
              linetype = "dashed", linewidth = 1) +
  labs( x = "Measured Adulteration (%)",
    y = "Predicted Adulteration (%)") +
  annotate("text",
    x = min(observed), y = max(predictions),
    label = paste("RMSEP:", round(RMSEP, 2),
      "\nR-squared:", round(R2, 3),
      "\nRPD:", round(RPD, 2)),
    hjust = 0, vjust = 1, size = 3, color = "black") +
  theme(
    axis.title = element_text(color = "black", size = 10),
    axis.text = element_text(color = "black"))+
  theme_gray()

```

```{r}
# Check residuals

residuals <- observed - predictions

# Plot Residuals
plot(predictions, residuals,
     xlab = "Predicted Values",
     ylab = "Residuals",
     main = "Residuals vs Predicted",
     pch = 16, col = "blue")
abline(h = 0, col = "red", lwd = 2)  
```


```{r}
hist(residuals, 
     main = "Histogram of Residuals",
     xlab = "Residuals", cex.main = 0.8,
     col = "lightblue", border = "black"); box()

```



## Conclusion

- PLS demonstrate good performance in predicting adulteration in ground coffee. The model achieves an RPD value of 3.2.