---
title: "Detection of Coffee Adulteration Using Hyperspectral Imaging and Random Forest: Spectra Preprocessed by MSC, Savitzky Golay and 2nd Derivative (MSC+SG+2nd Der.)"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: false
---

## Introduction

This project aims to distinguish authentic coffee from adulterated samples using Random Forest, a robust ensemble machine learning algorithm known for its effectiveness in handling high-dimensional and complex data. By analyzing spectral data, the model will focus on accurately classifying pure coffee and adulterated samples across varying levels of adulteration, ensuring reliable detection and minimizing classification errors.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class) 
  library(randomForest)
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(gridExtra)
  library(kernlab)}))
```

## Load and inspect data: msc_sg_2d spectra

```{r}
hsi_msc_sg_2d <- read.csv("coffee_ground_msc_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_2d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_msc_sg_2d$binary_class <- as.factor(hsi_msc_sg_2d$binary_class)
```

```{r}
# Get the X and y variables
X <- hsi_msc_sg_2d[,c(6:229)]
y <- hsi_msc_sg_2d[,2]
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue', freq = FALSE, 
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()
```

## Split samples into training and test sets

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_2d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)

# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Chek composition of the y variable for both train and test sets
table(train_data$binary_class)
table(test_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Balance the data using SMOTE

```{r}
# Due to data imbalance, we will use SMOTE to balance the data set

warning = FALSE  
  suppressWarnings(suppressMessages({
library(DMwR)  
}))


# Combine the dataset
smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)

# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

# Reduction Dimensional (PCA), Centering and Scaling Data

```{r}
# Reduction dimensionality to eliminate collinearity, centering and scaling

# Imabalanced Data
pca_Preprocess<- preProcess(train_data, method = c("center", "scale", "pca"))
data_trainPreprocessed <- predict(pca_Preprocess, train_data)
data_testPreprocessed <- predict(pca_Preprocess, test_data)

#SMOTE data
pca_Preprocess_smote<- preProcess(smote_data, method = c("center", "scale", "pca"))
data_trainPreprocessed_smote <- predict(pca_Preprocess_smote, smote_data)
data_testPreprocessed_smote <- predict(pca_Preprocess_smote, test_data)
```

## Set Up Training Parameters

```{r}
# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# OneSE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# Hyperparameter Tuning
# We will use tuneLength to automatically determine the optimal parameters 
#The number of trees will be set to 500 (default)
```

## Train the Models

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_rf_msc_sg_2d<-train(binary_class~., data = data_trainPreprocessed, 
                      method = 'rf', tuneLength = 10, 
                      trControl = control, metric = 'Accuracy') #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)

print(fit_rf_msc_sg_2d)
```

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

fit_rf_smote_msc_sg_2d<-train(binary_class~., data = data_trainPreprocessed_smote,
                            method ='rf',tuneLength = 10, 
                            trControl = control, metric = 'Accuracy')# SMOTE data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)

print(fit_rf_smote_msc_sg_2d)
```

## Save the models for future use

```{r}
saveRDS(fit_rf_msc_sg_2d, "fit_rf_msc_sg_2d.rds")
saveRDS(fit_rf_smote_msc_sg_2d, "fit_rf_smote_msc_sg_2d.rds")
```

## Plot the cross-validation results ----\> Model selection

```{r}
# Imbalanced Data
mtry_data <- fit_rf_msc_sg_2d$results
best_mtry = fit_rf_msc_sg_2d$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]
plot(mtry_data$mtry,mtry_data$Accuracy, type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     xlab = "Number of mtry", ylab = "Accuracy (%)",cex.lab = 0.9)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

```{r}
# Balanced Data
mtry_data <- fit_rf_smote_msc_sg_2d$results
best_mtry = fit_rf_smote_msc_sg_2d$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]
plot(mtry_data$mtry,mtry_data$Accuracy, type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     xlab = "Number of mtry", ylab = "Accuracy (%)",cex.lab = 0.9)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

## Display the cross-validation results

```{r}
print(paste('The optimal parametrers for training the rf model are mtry value of', fit_rf_msc_sg_2d$bestTune$mtry)) # unmodified data 

print(paste('The optimal parametrers for training the rf model are mtry value of', fit_rf_smote_msc_sg_2d$bestTune$mtry)) # SMOTE data 
```

\#

```{r}
# Imbalanced data
knitr::kable(fit_rf_msc_sg_2d$results)
```

```{r}
# Select the best results row 
selected_row <- fit_rf_msc_sg_2d$results %>% 
  filter(mtry==fit_rf_msc_sg_2d$bestTune$mtry)

knitr::kable(selected_row)
```

```{r}
# SMOTE Data
knitr::kable(fit_rf_smote_msc_sg_2d$results)

```

```{r}
# Select the best results row 
selected_row_smote <- fit_rf_smote_msc_sg_2d$results %>% 
  filter(mtry== fit_rf_smote_msc_sg_2d$bestTune$mtry)

knitr::kable(selected_row_smote)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_rf_msc_sg_2d$pred$pred,fit_rf_msc_sg_2d$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_rf_smote_msc_sg_2d$pred$pred,fit_rf_smote_msc_sg_2d$pred$obs) # SMOTE data
```

## Test the model performance on test data for unbalanced trained data

```{r}
# Predict on the test set 
set.seed(123)
test_rf_unmodified <- predict(fit_rf_msc_sg_2d,newdata=data_testPreprocessed)

# Get the confusion matrix
test_rf_unmodified_CMtrx <- confusionMatrix(test_rf_unmodified,
                                             data_testPreprocessed$binary_class)
test_rf_unmodified_CMtrx

knitr::kable(test_rf_unmodified_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_rf_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_rf <- test_rf_unmodified_CMtrx$table
TP <- cf_table_rf[1,1]
TN <- cf_table_rf[2,2] 
FP <- cf_table_rf[2,1] 
FN <- cf_table_rf[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Predict on the test set 
set.seed(123)
test_rf_smote <- predict(fit_rf_smote_msc_sg_2d,newdata=data_testPreprocessed_smote)

# Get the confusion matrix
test_rf_smote_CMtrx <- confusionMatrix(test_rf_smote,
                                             data_testPreprocessed_smote$binary_class)
test_rf_smote_CMtrx

knitr::kable(test_rf_smote_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_rf_smote_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_rf_smote <- test_rf_smote_CMtrx$table
TP <- cf_table_rf_smote[1,1]
TN <- cf_table_rf_smote[2,2] 
FP <- cf_table_rf_smote[2,1] 
FN <- cf_table_rf_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(rf = fit_rf_msc_sg_2d, rf_smote = fit_rf_smote_msc_sg_2d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
p1<-ggplot(data = as.data.frame(test_rf_unmodified_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix RF Unbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(data = as.data.frame(test_rf_smote_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "azure1", high = "turquoise", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix RF with SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

```{r}
p2<-ggplot(data = as.data.frame(test_rf_smote_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix RF with SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

p2
```


```{r}
plt <- as.data.frame(test_rf_smote_CMtrx$table)
plt$Reference <- factor(plt$Reference, levels=rev(levels(plt$Reference)))

ggplot(plt, aes(Prediction,Reference, fill= Freq)) +
        geom_tile(aes(fill = Freq), color = "black") + geom_text(aes(label=Freq)) +
        scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix Random Forest (Ground Coffee)')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

```



## ROC Curve

```{r}
#SMOTE data
roc_curve <- roc(data_testPreprocessed_smote$binary_class, as.numeric(test_rf_smote))
```

```{r}
auc_value <- auc(roc_curve)
plot(roc_curve, col = '#1f77b4', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)")
abline(a = 0, b = 1, lty = 2, col = "black")
legend("bottomright", legend = c("RF"), col = '#1f77b4', lwd = 2)
```

### Decision Boundary for RF Model

```{r}
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

train_subset <- train_merged_smote_Preprocessed %>% dplyr::select(PC1,PC2,binary_class)

control_1 <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE')

rf_grid_fit <- train(binary_class~., data = train_subset, 
                       method = 'rf',trControl=control_1,tuneLength=40,
                      metric = metric)
stopCluster(cl)
```


```{r}
# Create a grid of values
pc1_range <- seq(min(train_subset$PC1), max(train_subset$PC1), length.out = 100)
pc2_range <- seq(min(train_subset$PC2), max(train_subset$PC2), length.out = 100)

# Expand grid
grid_data <- expand.grid(PC1 = pc1_range, PC2 = pc2_range)

# Predict class probabilities for each point in the grid
grid_data$prediction <- predict(rf_grid_fit, newdata = grid_data)
grid_data$prediction <- as.factor(grid_data$prediction)
```


```{r}
library(ggplot2)
library(viridis)  # Load the viridis package

ggplot() +
  geom_tile(data = grid_data, aes(x = PC1, y = PC2, fill = prediction), alpha = 0.8) +
  geom_point(data = train_subset, aes(x = PC1, y = PC2, color = binary_class, shape = binary_class), size = 3) +
  labs(
    title = "Random Forest Decision Boundary", 
    x = "PC1", 
    y = "PC2",
    color = "Class",
    fill = "Decision Region",
    shape = "Class"
  ) +
  theme_bw() +
  scale_fill_viridis_d(option = "D", begin = 0.2, end = 0.8) +  # Viridis palette for fill
  scale_color_manual(values = c("adulterated" = "black", "pure_arabica" = "blue")) +
  scale_shape_manual(values = c("adulterated" = 17, "pure_arabica" = 1)) +   
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 12),  
    axis.title = element_text(size = 12, face = "bold"),  
    axis.text = element_text(size = 10),  
    legend.position = "bottom",  
    legend.direction = "horizontal",  
    legend.box = "horizontal",
    legend.spacing.x = unit(0.1, 'cm'),  
    legend.spacing.y = unit(0.1, 'cm'),
    legend.key.height = unit(0.2, "cm"),  
    legend.key.width = unit(0.3, "cm"),   
    legend.background = element_blank(),  
    legend.title = element_text(size = 8, face = "bold"),  
    legend.text = element_text(size = 9)
  )

```

### Misclassifications 

```{r}
adult_test <- hsi_msc_sg_2d$adult_percent[!sample]
predicted <- test_rf_smote
observed <- y_test
percent_aldult <- adult_test

df_pred <- as.data.frame(cbind(observed, predicted, percent_aldult))
head(df_pred)

misclassif <- df_pred %>%
  filter(observed != predicted)

misclassif

```

## Conclusion

-   RF model based on SNV+SG+1st derivative achieve exceptional performance. The model achieve MCC > 0.9.
