---
title: "Feature Optimization for Coffee Adulteration Detection Using Hyperspectral Imaging and Ensemble Machine Learning: Integrating Boruta with SNV+SG+2nd Derivative Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "sky"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
  cache: true
---

## Introduction

This project explores the effectiveness of ensemble learning methods in distinguishing authentic coffee from adulterated samples using spectral data derived from hyperspectral imaging (HSI). Leveraging Boruta for feature selection, the study identifies the most relevant spectral features, significantly reducing dimensionality while maintaining high model accuracy. By addressing the curse of dimensionality, this approach ensures the ensemble models focus on the most informative wavelengths, enhancing their predictive performance.

The ensemble learning framework combines predictions from diverse base models, including Random Forest (RF), Support Vector Machines (SVM), Artificial Neural Networks (ANN), Linear Discriminant Analysis (LDA), and k-Nearest Neighbors (kNN), to improve robustness and classification accuracy. By employing stacking and voting approaches, the study integrates the strengths of these algorithms, achieving better generalization across varying levels of coffee adulteration.The ensemble’s performance is rigorously evaluated using 10-fold repeated cross-validation, ensuring reliability and minimizing the risk of overfitting.

The study focuses on accurately classifying pure and adulterated coffee across a range of adulteration levels (e.g., 1%, 5%, 10%, and higher concentrations). Furthermore, it critically examines the impact of feature selection and ensemble learning on computational efficiency, sensitivity, and specificity, offering a robust framework for applying advanced machine learning techniques in food authenticity and fraud detection.
## Setup and Load Packages

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis) 
  library(NeuralNetTools)
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(caretEnsemble)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(Boruta)
  library(ggthemes)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: snv_sg_2d spectra

hsi_snv_sg_2d <- read.csv("coffee_instant_snv_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_snv_sg_2d)

# Check a few column names
col_check <- colnames(hsi_snv_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6 -------> 229

# Convert the class to a factor
hsi_snv_sg_2d$binary_class <- as.factor(hsi_snv_sg_2d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_snv_sg_2d[,c(6:229)]
y <- hsi_snv_sg_2d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
# Create histogram of spectral intensities with improved formatting
hist(as.matrix(X), 
     col = "lightblue",
     border = "white",  
     main = "Distribution of Spectral Intensity Values",  
     xlab = "Spectral Intensity (AU)",  
     ylab = "Frequency",  
     las = 1,  
     cex.main = 1.2,  
     cex.lab = 1.1,  
     breaks = 30)  

# Add grid lines 
grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")

# Add box around plot 
box(lwd = 2)                              
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_snv_sg_2d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("slateblue", "sienna")) +  # Specify your custom colors here
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +  # Ensure legend names are retained
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5, size = 10))
```

## **Balance the data using SMOTE**

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```
## Run Boruta Algorithm for Optimum Wavelength Selection

```{r}

  # Set seed for reproducibility 
  set.seed(123)

  # Set the dataset name dynamically
  dataset_name <- "snv_sg_2d"  

  # Define paths for the execution flag and Boruta result
  flag_file <- paste0("boruta_executed_flag_", dataset_name, ".RData")
  result_file <- paste0("boruta_result_", dataset_name, ".RData")

  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

if (!file.exists(result_file)) {
  message(paste("Starting Boruta execution for dataset:", dataset_name))
  
  # Start the timer
  start_time <- Sys.time()

  # Run Boruta on the train dataset
  boruta_result <- Boruta(formula = binary_class ~ .,  
                        data = train_data, 
                        doTrace = 2,
                        maxRuns = 100)                   
 
  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the Boruta model:', end_time - start_time))  
  
  # Save Boruta results and execution flag
  boruta_executed <- TRUE
  save(boruta_executed, file = flag_file)
  save(boruta_result, file = result_file)
 

  message(paste("Boruta execution completed for dataset:", dataset_name, "and results saved."))

} else {
  # Load previously saved results
  message(paste("Result file found. Loading saved Boruta results for dataset:", dataset_name))
  load(result_file)
  load(flag_file)
  message(paste("Boruta already executed for dataset:", dataset_name))
}

```


```{r}
# Print Boruta summary
print(boruta_result)
```

```{r}
# Get selected features
final_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
final_features
```


```{r}
# Plot Boruta results
plot(boruta_result, xlab = "", las = 2, 
     cex.main = 0.8, cex.axis = 0.6, 
     main = paste("Boruta Feature Selection"))
```

```{r}
print(paste('The number of selected variables are', length(final_features)))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, final_features, drop = FALSE]  
X_test_selected <- X_test[, final_features, drop = FALSE]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(X_train_selected, binary_class = y_train_selected)
test_merged <- cbind(X_test_selected, binary_class = y_test_selected)

dim(train_merged)
dim(test_merged)
table(train_merged$binary_class)
table(test_merged$binary_class)

```

## Create a Data frame with the Selected Features for SMOTE Data

```{r}
X_train_selected_smote <- X_train_smote[,final_features, drop = FALSE]
X_test_selected_smote <- X_test[, final_features, drop = FALSE] 

# Keep y_train and y_test as-is 
y_train_selected_smote <- y_train
y_test_selected_smote <- y_test

#Merge data_frame
train_merged_smote <- cbind(X_train_selected_smote, binary_class = smote_data$binary_class)
test_merged_smote <- cbind(X_test_selected_smote, binary_class = y_test_selected_smote)

dim(train_merged_smote)
dim(test_merged_smote)
table(train_merged_smote$binary_class)
table(test_merged_smote$binary_class)
```

## Check correlations by HeatMaps

```{r}

n_features <- length(final_features)

# Calculate the correlation matrix 

#Imbalanced Data

#Calculate correlations
corr <- cor(train_merged[, -(n_features+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("blue", "white", "red"))(5), 
       title = "Correlation", cex = 0.8) 

```

```{r}
# SMOTE Data
par(cex.main = 0.6)
corr <- cor(train_merged_smote[, -(n_features+1)])
# Create heat map
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("#440154", "#21908C", "gold"))(50),
        main = 'Correlation Heatmap',cex.main = 0.6)  

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("#440154", "#21908C", "gold"))(5), 
       title = "Correlation", 
       cex = 0.8) 
```

-   The correlation heat maps reveal that some variables remain highly correlated. To address this issue and mitigate multicollinearity, we will employ Principal Component Analysis (PCA)

## PreProcess by PCA, Centering and Scaling Data

```{r}

# Imbalanced Data
pca_Preprocess <- preProcess(train_merged, method = c("center", "scale", "pca"))
train_mergedPreprocessed <- predict(pca_Preprocess, train_merged) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,test_merged) # apply on test data

# SMOTE Data
pca_Preprocess_smote <- preProcess(train_merged_smote,method = c("center", "scale", "pca"))
train_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,train_merged_smote)
test_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,test_merged_smote)

# Check dimensions of the preprocessed data to ensure consistency

dim(train_mergedPreprocessed)
dim(test_mergedPreprocessed)

dim(train_merged_smote_Preprocessed)
dim(test_merged_smote_Preprocessed)
```
## Set Up Training Parameters

```{r}

# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# One SE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# We will train all the models: kNN, NNET, SVM, LDA and RF as base learners. Then the predictions from this model will be trained by RF as a  meta-learner before testing on an external validation set.

# Hyper-parameter Tuning: We will train the model with their own set of hyper parameter settings.

# LDA model: No hyper parameter

# k-NN
grid_knn <- expand.grid(.k = seq(3,30, by = 2))

# SVM. We will use a tuneLength of 10. This will automatically select optimal Cost (C) and sigma parameter for the RBF kernel 

# RF: We will also use a tuneLength of 10. This selects the optimal mtry/number of splits

# NNET model. Selects the optimal number of neurons and decay value  (regularization parameter)
grid_nnet <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

## Train the Models

```{r cache=TRUE}

# Unmodified data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_snv_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'lda',trControl = control, metric = 'Accuracy')

# 2. Train the kNN model
fit_knn_model_snv_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'knn',trControl = control, 
                       tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_snv_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'svmRadial',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_snv_sg_2d <- train(binary_class ~., data = train_mergedPreprocessed,
                       method = 'rf',trControl = control, 
                       tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_snv_sg_2d<-train(binary_class ~., data = train_mergedPreprocessed, 
                      method = 'nnet', tuneGrid = grid_nnet,
                      trControl = control, metric = 'Accuracy',
                      trace = FALSE, MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# After training the models, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list = list(lda = fit_lda_model_snv_sg_2d, 
                   knn = fit_knn_model_snv_sg_2d,
                   svm = fit_svm_model_snv_sg_2d,
                   rf = fit_rf_model_snv_sg_2d,
                   nnet = fit_nnet_snv_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_snv_sg_2d <- caretStack(models_list, method = 'rf', metric = metric,
                                       trControl = control,tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```
```{r}
# Train the models on SMOTE balanced data

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)

# 1. Train the LDA model
fit_lda_model_smote_snv_sg_2d <- train(binary_class ~.,
                                       data = train_merged_smote_Preprocessed,
                                       method = 'lda',trControl = control, 
                                       metric = 'Accuracy')

# 2. Train the kNN model
fit_knn_model_smote_snv_sg_2d <- train(binary_class ~., 
                                 data = train_merged_smote_Preprocessed,
                                 method = 'knn',trControl = control, 
                                 tuneGrid = grid_knn, metric = 'Accuracy')

# 3. Train the SVM model
fit_svm_model_smote_snv_sg_2d <- train(binary_class ~., 
                                       data = train_merged_smote_Preprocessed,
                                       method = 'svmRadial',trControl = control,
                                       tuneLength = 10, metric = 'Accuracy')

# 4. Train the RF model
fit_rf_model_smote_snv_sg_2d <- train(binary_class ~., 
                                data = train_merged_smote_Preprocessed,
                                method = 'rf',trControl = control,
                                tuneLength = 10, metric = 'Accuracy')

# 5. Train the NNET model

fit_nnet_smote_snv_sg_2d<-train(binary_class ~., 
                          data = train_merged_smote_Preprocessed,
                          method = 'nnet', tuneGrid = grid_nnet,
                          trControl = control, metric = 'Accuracy',
                          trace = FALSE, MaxNWts = 1000, maxit = 300) 
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```



```{r}
# Similarly, we will stack/combine them in an ensemble

# Combine models in an ensemble

models_list_smote = list(lda = fit_lda_model_smote_snv_sg_2d, 
                         knn = fit_knn_model_smote_snv_sg_2d,
                         svm = fit_svm_model_smote_snv_sg_2d,
                         rf = fit_rf_model_smote_snv_sg_2d,
                         nnet = fit_nnet_smote_snv_sg_2d)

# Then train using RF as a meta-learner

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()


fit_stack_ensemble_smote_snv_sg_2d <- caretStack(models_list_smote, 
                                             method = 'rf', 
                                             metric = metric,
                                             trControl = control,
                                             tuneLength = 10)

#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
print(fit_stack_ensemble_snv_sg_2d) # Model on imbalanced data
```

```{r}
print(fit_stack_ensemble_smote_snv_sg_2d) # SMOTE model
```


## Save the models for future use

```{r}
saveRDS(fit_stack_ensemble_snv_sg_2d, 
        "fit_stack_ensemble_snv_sg_2d.rds") # for unmodified data

saveRDS(fit_stack_ensemble_smote_snv_sg_2d, 
        "fit_stack_ensemble_smote_snv_sg_2d.rds") # SMOTE data
```

## Display the optimal parameters for the ensemble meta learner

```{r}
# Used 500 trees as a default
print(paste('The optimal parametrers for training the meta-learner rf model (on unmodified data) are mtry value of', 
            fit_stack_ensemble_snv_sg_2d$ens_model$bestTune)) # unmodified data 

print(paste('The optimal parametrers for training the meta-learner rf model (on data balanced by SMOTE) are  mtry value of', fit_stack_ensemble_smote_snv_sg_2d$ens_model$bestTune)) # unmodified data  # SMOTE data 
```

## Display the cross-validation results in Tables

```{r warning=FALSE}
# Imbalanced data
kable(fit_stack_ensemble_snv_sg_2d$ens_model$results, 
      caption = "Cross-Validation Results on Imbalanced Data") # Imbalanced Data
```


```{r warning=FALSE}
# SMOTE data
kable(fit_stack_ensemble_smote_snv_sg_2d$ens_model$results,
      caption = "Cross-Validation Results on SMOTE Data") # Balanced Data
```

```{r}
# Select the best results row for unmodified data
selected_row <- fit_stack_ensemble_snv_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_snv_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)
```

```{r}
# Select the best results row for SMOTE Data
selected_row <- fit_stack_ensemble_smote_snv_sg_2d$ens_model$results %>% 
  filter(mtry == fit_stack_ensemble_smote_snv_sg_2d$ens_model$bestTune$mtry)

kable(selected_row)

```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_stack_ensemble_snv_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_snv_sg_2d$ens_model$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_stack_ensemble_smote_snv_sg_2d$ens_model$pred$pred,
                fit_stack_ensemble_smote_snv_sg_2d$ens_model$pred$obs) # SMOTE data
```

Visualize Hyperparameter Tuning for Optimal Model Selection

```{r}

# Imbalanced Data

mtry_data <- fit_stack_ensemble_snv_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_snv_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

```{r}

# SMOTE Data

mtry_data <- fit_stack_ensemble_smote_snv_sg_2d$ens_model$results
best_mtry = fit_stack_ensemble_smote_snv_sg_2d$ens_model$bestTune$mtry
y_Acc = mtry_data$Accuracy[mtry_data$mtry == best_mtry]

#Plot the CV results

plot(mtry_data$mtry,mtry_data$Accuracy, 
     type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     ylim = c(0.6, 1),
     xlab = "Number of mtry", 
     ylab = "Accuracy (%)",cex.lab = 0.9, main = 'Optimal selection of mtry',
     cex.main = 0.8)
# Add line from x-axis to the specific point
segments(x0 = best_mtry, y0 = 0, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_mtry, y1 = y_Acc, col = "black", lty = 2)
```

## Visualize the Out of Bag Errors (OOB)

```{r}

# Unmodified data

# Plot the OOB error from the ensemble model
plot(fit_stack_ensemble_snv_sg_2d$ens_model$finalModel, 
     main = 'OOB Error Rate', 
     cex.main = 0.8,
     cex.axis = 0.8,
     col = c('blue', 'red', 'black'),
     lwd = 1.5)

# Add legend
legend("topright",
       legend = c("OOB Error", "Adulterated", "Arabica"),
       col = c('blue', 'red', 'black'),
       lty = c(1,2,2),
       lwd = 1.5,
       cex = 0.8)
```

```{r}
# SMOTE Data 

# Plot the OOB error from the ensemble model
plot(fit_stack_ensemble_smote_snv_sg_2d$ens_model$finalModel, 
     main = 'OOB Error Rate', 
     cex.main = 0.8,
     cex.axis = 0.8,
     col = c('blue', 'red', 'black'),
     lwd = 1.5)

# Add legend
legend("topright",
       legend = c("OOB Error", "Adulterated", "Arabica"),
       col = c('blue', 'red', 'black'),
       lty = c(1,2,2),
       lwd = 1.5,
       cex = 0.8)
```

## Test the model peformance on test data for unbalanced trained data

```{r}
# Make predictions with the ensemble model: Imbalanced Data

test_ensemble_unmodified <- predict(fit_stack_ensemble_snv_sg_2d,
                                newdata=test_mergedPreprocessed)

# Apply threshold to probabilities
class_predictions_unmodified <- ifelse(test_ensemble_unmodified[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_unmodified <- factor(class_predictions_unmodified, 
                                       levels = c("adulterated", "pure_arabica"))

reference_unmodified <- test_mergedPreprocessed$binary_class

# Get the confusion matrix

test_ensemble_unmodified_cfmatrix <- confusionMatrix(class_predictions_unmodified,
                                          reference_unmodified)
test_ensemble_unmodified_cfmatrix

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_unmodified_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_unmodified_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Make predictions with the ensemble model: SMOTE Data

test_ensemble_smote <- predict(fit_stack_ensemble_smote_snv_sg_2d,
                                newdata=test_merged_smote_Preprocessed)

# Apply threshold to probabilities
class_predictions_smote <- ifelse(test_ensemble_smote[, "adulterated"] >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

# Convert to a factor
class_predictions_smote <- factor(class_predictions_smote, 
                                       levels = c("adulterated", "pure_arabica"))

reference_smote <- test_merged_smote_Preprocessed$binary_class

# Get the confusion matrix

test_ensemble_smote_cfmatrix <- confusionMatrix(class_predictions_smote,
                                          reference_smote)
test_ensemble_smote_cfmatrix

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$byClass)

# View the results as knitr table
kable(test_ensemble_smote_cfmatrix$table)


# Extracting the components of the confusion matrix
cf_table_nnet <- test_ensemble_smote_cfmatrix$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(ensemble_1 = fit_stack_ensemble_snv_sg_2d$ens_model, 
                          ensemble_smote = fit_stack_ensemble_smote_snv_sg_2d$ens_model))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert re-sampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results, col = 'red')

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_ensemble_unmodified_cfmatrix$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_smote <- as.data.frame(test_ensemble_smote_cfmatrix$table)
plt_smote$Reference <- factor(plt_smote$Reference, levels=rev(levels(plt_smote$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: ensemble Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_smote, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: ensemble SMOTE Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(test_merged_smote_Preprocessed$binary_class,
                 as.numeric(class_predictions_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```



```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("ensemble"), col = 'blue', lwd = 3)
```

### Misclassifications

```{r}
## Imbalanced
observed <- y_test
predicted <- test_ensemble_unmodified
perc <- hsi_snv_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))

imbal_class$adulterated<- ifelse(imbal_class$adulterated >= 0.5, 
                             "adulterated", 
                             "pure_arabica")

imbal_class$pure_arabica<- ifelse(imbal_class$pure_arabica >= 0.5, 
                             "pure_arabica",
                              "adulterated") 
                             

misclas_imbal <- imbal_class %>% dplyr::filter(observed!=adulterated & observed!=pure_arabica)
misclas_imbal

```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# SMOTE Data
observed <- y_test
predicted <- test_ensemble_smote
perc <- hsi_snv_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

bal_class <- as.data.frame(cbind(observed,predicted,perc))

bal_class$adulterated<- ifelse(bal_class$adulterated >= 0.5, 
                             "adulterated", 
                             "pure_arabica")
bal_class$pure_arabica<- ifelse(bal_class$pure_arabica >= 0.5, 
                             "pure_arabica",
                              "adulterated") 

misclas_smote <- bal_class %>% filter(observed!=adulterated & observed!=pure_arabica)
misclas_smote
```

```{r}
table(misclas_smote$perc_adulter)
```

