---
title: "Feature Optimization for Coffee Adulteration Detection Using Hyperspectral Imaging and Artificial Neural Networks: RFE and GA with MSC+SG+2nd Derivative Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "sky"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
execute: 
  cache: true
---

## Introduction

This project leverages Artificial Neural Networks (ANN) to evaluate their effectiveness in distinguishing authentic coffee from adulterated samples using spectral data. By incorporating feature selection techniques such as Recursive Feature Elimination (RFE) and Genetic Algorithms (GA), the study aims to identify the most relevant spectral features, reducing dimensionality while maintaining model accuracy. ANN's ability to model complex, non-linear relationships makes it ideal for analyzing high-dimensional spectroscopic datasets. The study evaluates the model's performance in classifying pure and adulterated coffee across varying adulteration levels.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis) 
  library(NeuralNetTools)
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(ggthemes)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: msc_sg_2d spectra

hsi_msc_sg_2d <- read.csv("coffee_instant_msc_sg_2d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_2d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_2d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_msc_sg_2d$binary_class <- as.factor(hsi_msc_sg_2d$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_msc_sg_2d[,c(6:229)]
y <- hsi_msc_sg_2d[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
hist(as.matrix(X), 
     col = "blue",
     border = "white",  
     main = "Distribution of Spectral Intensity Values",  
     xlab = "Spectral Intensity (AU)",  
     ylab = "Frequency",  
     las = 1,  
     cex.main = 1.2,  
     cex.lab = 1.1,  
     breaks = 30)  

# Add grid lines 
grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")

# Add box around plot 
box(lwd = 2)
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_2d$cal_val == 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#1f77b4", "#ff7f0e")) +  # Specify your custom colors here
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +  # Ensure legend names are retained
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Balance the data using SMOTE

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the X and y for the over-sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("#17becf", "#ff7f0e")) +  # Blue and Orange
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count",
       fill = "Class") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Define the Fitness Function for GA

- This function evaluates subsets of features selected by GA using RFE with a Random Forest (RF) model.

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}

# Fitness Function
fitness_ga_rfe <- function(subset) {
  # Identify selected features based on the binary subset
  selected_features <- which(subset == 1)
  
  # Penalize empty subsets to avoid errors
  if (length(selected_features) == 0) return(Inf)
  
  # Subset the training feature matrix to selected features
  X_selected <- X_train[, selected_features]
  
  # Define control for rfe
  control <- rfeControl(functions = rfFuncs,  # Use Random Forest for rfe
                        method = "cv",       # Cross-validation
                        number = 10)         # 10-fold CV
  
  # Apply rfe to the selected subset of features
  rfe_result <- rfe(X_selected, y_train, rfeControl = control)
  
  # Return the negative accuracy as the fitness score (to maximize accuracy)
  return(-rfe_result$results$Accuracy[which.max(rfe_result$results$Accuracy)])
}
```

## Run the Genetic Algorithm

```{r}

# Set seed for reproducibility 
set.seed(123)

num_features <- ncol(X_train) # Number of features
dataset_name <- "msc_sg_2d"   # Set the dataset name dynamically

# Define paths for the execution flag and GA result
flag_file <- paste0("ga_executed_flag_", dataset_name, ".RData")
result_file <- paste0("ga_result_", dataset_name, ".RData")

# Check if the result file exists
if (!file.exists(result_file)) {  
  message("No result file found. Running the Genetic Algorithm (GA).")
  
  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

  # Register start time
  start_time <- Sys.time()

  # Run the Genetic Algorithm
  ga_result <- ga(
    type = "binary",                  
    fitness = fitness_ga_rfe,       
    nBits = num_features,             # Number of features
    popSize = 20,                     # Population size
    maxiter = 50,                     # Maximum iterations
    run = 10,                         # Convergence stopping criterion
    names = colnames(train_data[, -ncol(train_data)]),  # Feature names
    pcrossover = 0.8,                 # Probability of crossover
    pmutation = 0.1,                  # Probability of mutation
    maxFitness = -0.9,                # Stop when fitness reaches target
    parallel = TRUE,                  # Enable parallel processing
    monitor = gaMonitor,              # Track progress
    seed = 123                        # Reproducibility
  )

  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the GA model:', end_time - start_time))

  # Save the GA result and execution flag
  ga_executed <- TRUE
  save(ga_executed, ga_result, file = result_file)  # Save both flag and results
  
  message("GA execution completed and results saved.")
  
} else {
  # Load previously saved results
  message("Result file found. Loading saved GA results.")
  load(result_file)
  message(paste("GA already executed for dataset:", dataset_name))
}

```

## Visualization Plot

```{r}

par (cex = 0.8)
plot(ga_result, main = "GA Optimization Progress")
par(cex = 1)

```

## Analyze GA results

```{r}
summary(ga_result)
```

## Post-Selection

```{r}
# Get the best feature indices selected by GA
best_features <- which(ga_result@solution == 1)

# Remove indices beyond available columns
best_features <- best_features[best_features <= ncol(train_data)]

# Get selected feature names
selected_feature_names <- colnames(train_data)[best_features]

print(selected_feature_names)
```


```{r}
print(paste('The number of selected variables are', length(selected_feature_names)))
n_selected <- as.numeric(length(selected_feature_names))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, selected_feature_names, drop = FALSE]  
X_test_selected <- X_test[, selected_feature_names, drop = FALSE]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(X_train_selected, binary_class = y_train_selected)
test_merged <- cbind(X_test_selected, binary_class = y_test_selected)

dim(train_merged)
dim(test_merged)
table(train_merged$binary_class)
table(test_merged$binary_class)
```

## Create a Data frame with the Selected Features for SMOTE Data

```{r}
X_train_selected_smote <- X_train_smote[,selected_feature_names, drop = FALSE]
X_test_selected_smote <- X_test[, selected_feature_names, drop = FALSE] 

# Keep y_train and y_test as-is 
y_train_selected_smote <- y_train
y_test_selected_smote <- y_test

#Merge data_frame
train_merged_smote <- cbind(X_train_selected_smote, binary_class = smote_data$binary_class)
test_merged_smote <- cbind(X_test_selected_smote, binary_class = y_test_selected_smote)

dim(train_merged_smote)
dim(test_merged_smote)
table(train_merged_smote$binary_class)
table(test_merged_smote$binary_class)
```

## Check correlations by HeatMaps

```{r}
# Calculate the correlation matrix 

#Imbalanced Data

#Calculate correlations
corr <- cor(train_merged[, -(n_selected+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("blue", "white", "red"))(5), 
       title = "Correlation", cex = 0.8) 

```

```{r}
# SMOTE Data
par(cex.main = 0.6)
corr <- cor(train_merged_smote[, -(n_selected+1)])
# Create heat map
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("#440154", "#21908C", "gold"))(50),
        main = 'Correlation Heatmap',cex.main = 0.6)  

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("#440154", "#21908C", "gold"))(5), 
       title = "Correlation", 
       cex = 0.8) 
```

-   The correlation heat maps reveal that some variables remain highly correlated. To address this issue and mitigate multicollinearity, we will employ Principal Component Analysis (PCA). We will then proceed to train our models with features generated by PCA. This aims to speed up processing time.

## PreProcess by PCA, Centering and Scaling Data

```{r}

# Imbalanced Data
pca_Preprocess <- preProcess(train_merged, method = c("center", "scale", "pca"))
train_mergedPreprocessed <- predict(pca_Preprocess, train_merged) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,test_merged) # apply on test data

# SMOTE Data
pca_Preprocess_smote <- preProcess(train_merged_smote,method = c("center", "scale", "pca"))
train_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,train_merged_smote)
test_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,test_merged_smote)

# Check dimensions of the preprocessed data to ensure consistency

dim(train_mergedPreprocessed)
dim(test_mergedPreprocessed)

dim(train_merged_smote_Preprocessed)
dim(test_merged_smote_Preprocessed)
```

## Set Up Training Parameters

```{r}

# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# One SE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# Hyperparameter Tuning
# Set tune grid to find optimal value of size and decay on cross-validation
# Size is the number of number of units (neurons) in the hidden layer of the neural network
# Decay is the regularization parameter; adds penalty to the loss function

grid <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

## Train the Models

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_nnet_msc_sg_2d<-train(binary_class ~., data = train_mergedPreprocessed, 
                      method = 'nnet', tuneGrid = grid, 
                      trControl = control, metric = 'Accuracy',
                      trace = FALSE, MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
print(fit_nnet_msc_sg_2d) # Model on imbalanced data
```


```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

fit_nnet_smote_msc_sg_2d<-train(binary_class~., data = train_merged_smote_Preprocessed,
                            method ='nnet', tuneGrid = grid,
                            trControl = control, metric = 'Accuracy',
                            trace = FALSE, MaxNWts = 1000, maxit = 300)# SMOTE data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
print(fit_nnet_smote_msc_sg_2d)
```

## Save the models for future use

```{r}
saveRDS(fit_nnet_msc_sg_2d, "fit_nnet_msc_sg_2d.rds")
saveRDS(fit_nnet_smote_msc_sg_2d, "fit_nnet_smote_msc_sg_2d.rds")
```

## Display the cross-validation results

```{r}
print(paste('The optimal parametrers for training the nnet model are', round(as.numeric(fit_nnet_msc_sg_2d$bestTune$size),2),'neurons','and decay value of',fit_nnet_msc_sg_2d$bestTune$decay)) # unmodified data 

print(paste('The optimal parametrers for training the nnet model are', round(as.numeric(fit_nnet_smote_msc_sg_2d$bestTune$size),2),'neurons','and decay value of',fit_nnet_smote_msc_sg_2d$bestTune$decay)) # SMOTE data 
```

```{r}
# Imbalanced data
kable(fit_nnet_msc_sg_2d$results, 
      caption = "Cross-Validation Results on Imbalanced Data")
```

```{r}
# Select the best results row 
selected_row <- fit_nnet_msc_sg_2d$results %>% 
  filter(size==fit_nnet_msc_sg_2d$bestTune$size & 
           decay ==fit_nnet_msc_sg_2d$bestTune$decay)

kable(selected_row)
```

```{r}
# SMOTE Data
kable(fit_nnet_smote_msc_sg_2d$results, 
      caption = "Cross-Validation Results on SMOTE Data")

```

```{r}
# Select the best results row 
selected_row_smote <- fit_nnet_smote_msc_sg_2d$results %>% 
  filter(size==fit_nnet_smote_msc_sg_2d$bestTune$size & 
           decay ==fit_nnet_msc_sg_2d$bestTune$decay)

knitr::kable(selected_row_smote)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_nnet_msc_sg_2d$pred$pred,
                fit_nnet_msc_sg_2d$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_nnet_smote_msc_sg_2d$pred$pred,
                fit_nnet_smote_msc_sg_2d$pred$obs) # SMOTE data
```

Visualize Hyperparameter Tuning for Optimal Model Selection

```{r warning=FALSE}
# Imbalanced Data

# Get the best accuracy based on best size and decay
best_size <- fit_nnet_msc_sg_2d$bestTune$size
best_decay <- fit_nnet_msc_sg_2d$bestTune$decay
best_accuracy <- fit_nnet_msc_sg_2d$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_msc_sg_2d, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

```{r warning=FALSE}
# SMOTE data
# Get the best accuracy based on best size and decay
best_size <- fit_nnet_smote_msc_sg_2d$bestTune$size
best_decay <- fit_nnet_smote_msc_sg_2d$bestTune$decay
best_accuracy <- fit_nnet_smote_msc_sg_2d$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_smote_msc_sg_2d, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

## Visualize the NNET Model Architecture

```{r}
# Unmodified data
plotnet(fit_nnet_msc_sg_2d$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
```

```{r}
#SMOTE data
plotnet(fit_nnet_smote_msc_sg_2d$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
```

## Test the model peformance on test data for unbalanced trained data

```{r}
# Predict on the test set 
set.seed(123)
test_nnet_unmodified <- predict(fit_nnet_msc_sg_2d,
                                newdata=test_mergedPreprocessed)

# Get the confusion matrix
test_nnet_unmodified_CMtrx <- confusionMatrix(test_nnet_unmodified,
                                             test_mergedPreprocessed$binary_class)

test_nnet_unmodified_CMtrx # print the confusion matrix table

kable(test_nnet_unmodified_CMtrx$byClass) # Tabulate correct and incorrect classifications

# View the results as knitr table
kable(test_nnet_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient (MCC)

# Extracting the components of the confusion matrix
cf_table_nnet <- test_nnet_unmodified_CMtrx$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Predict on the test set 
set.seed(123)
test_nnet_smote <- predict(fit_nnet_smote_msc_sg_2d,
                           newdata=test_merged_smote_Preprocessed)

# Get the confusion matrix
test_nnet_smote_CMtrx <- confusionMatrix(test_nnet_smote,
                                             test_merged_smote_Preprocessed$binary_class)
test_nnet_smote_CMtrx # print the results 

kable(test_nnet_smote_CMtrx$byClass)

#View the results as knitr table
kable(test_nnet_smote_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_nnet_smote <- test_nnet_smote_CMtrx$table
TP <- cf_table_nnet_smote[1,1]
TN <- cf_table_nnet_smote[2,2] 
FP <- cf_table_nnet_smote[2,1] 
FN <- cf_table_nnet_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model performance
results <- resamples(list(nnet = fit_nnet_msc_sg_2d, nnet_smote = fit_nnet_smote_msc_sg_2d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert re-sampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_nnet_unmodified_CMtrx$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_smote <- as.data.frame(test_nnet_smote_CMtrx$table)
plt_smote$Reference <- factor(plt_smote$Reference, levels=rev(levels(plt_smote$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: nnet Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_smote, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: nnet SMOTE Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(test_merged_smote_Preprocessed$binary_class,
                 as.numeric(test_nnet_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("nnet"), col = 'blue', lwd = 3)
```

### Misclassifications

```{r}
## Imbalanced
observed <- y_test
predicted <- test_nnet_unmodified
perc <- hsi_msc_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_imbal <- imbal_class %>% filter(observed!=predicted)
misclas_imbal
```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# SMOTE Data
observed <- y_test
predicted <- test_nnet_smote
perc <- hsi_msc_sg_2d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

bal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_smote <- bal_class %>% filter(observed!=predicted)
misclas_smote
```

```{r}
table(misclas_smote$perc_adulter)
```