---
title: "Boruta Feature Selection and Neural Networks for Detecting Adulteration in Ground Coffee: Raw Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: "darkly"
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

Coffee adulteration, such as blending premium Arabica with lower-cost Robusta, undermines quality and authenticity, posing significant challenges for consumers, producers, and regulators. High-dimensional spectroscopic data provides valuable insights for detecting such fraud, but the sheer number of variables requires effective feature selection techniques to isolate the most relevant wavelengths.

This study leverages the Boruta algorithm, a robust feature selection method that identifies all relevant variables by iteratively comparing the importance of original features with randomized counterparts using a Random Forest framework. Unlike traditional methods, Boruta not only ensures the inclusion of critical spectral features but also prevents the selection of irrelevant or redundant variables, which is essential for improving model interpretability and accuracy.

By integrating Boruta with machine learning models, this study demonstrates an efficient and reliable approach to classify pure and adulterated coffee samples. The results highlight the potential of this method for detecting and quantifying coffee adulteration, ultimately supporting efforts to maintain product quality and authenticity in the coffee industry.


## Setup and Load Packages

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(GA)
  library(Boruta)
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls) 
  library(ggthemes)
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(DMwR)
  library(e1071) 
  library(knitr)
  library(pROC)
  library(gridExtra)
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(factoextra)
  library(kernlab)}))
```

## Prepare Data

```{r}
# Load data: raw spectra

hsi_raw <- read.csv("coffee_data_pycleaned.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_raw)

# Check a few column names
col_check <- colnames(hsi_raw[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_raw$binary_class <- as.factor(hsi_raw$binary_class)
```

## Get the X and y variables

```{r}
X <- hsi_raw[,c(6:229)]
y <- hsi_raw[,2]

class(y) #Ensure y (target variable) is a factor for classification
```

## Visualize X

```{r}
hist(as.matrix(X), col = 'lightblue',
     main = 'Histogram of Spectral Intensity Values', xlab = 'Intensity'); box()                                    
```

## Split the Data (Train_Test Split)

-   We want to select features based on training data only to avoid any data leakage that could inflate model performance and make the model less generalizable to unseen data.

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_raw$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)
```

```{r}
# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## **Balance the data using SMOTE**

```{r}

# Due to data imbalance, we will use SMOTE to balance the data set

smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)
```

```{r}
# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```
## Run Boruta Algorithm for Optimum Wavelength Selection

```{r}

  # Set seed for reproducibility 
  set.seed(123)

  # Set the dataset name dynamically
  dataset_name <- "raw"  

  # Define paths for the execution flag and Boruta result
  flag_file <- paste0("boruta_executed_flag_", dataset_name, ".RData")
  result_file <- paste0("boruta_result_", dataset_name, ".RData")

  # Register cluster for caret to train the models in parallel
  cl <- makeCluster(6, type = "SOCK")
  suppressWarnings(suppressMessages(registerDoSNOW(cl)))

if (!file.exists(result_file)) {
  message(paste("Starting Boruta execution for dataset:", dataset_name))
  
  # Start the timer
  start_time <- Sys.time()

  # Run Boruta on the train dataset
  boruta_result <- Boruta(formula = binary_class ~ .,  
                        data = train_data, 
                        doTrace = 2,
                        maxRuns = 100)                   
 
  # Stop the cluster
  stopCluster(cl)

  # Register end time
  end_time <- Sys.time()
  print(paste('Time to run the Boruta model:', end_time - start_time))  
  
  # Save Boruta results and execution flag
  boruta_executed <- TRUE
  save(boruta_executed, file = flag_file)
  save(boruta_result, file = result_file)
 

  message(paste("Boruta execution completed for dataset:", dataset_name, "and results saved."))

} else {
  # Load previously saved results
  message(paste("Result file found. Loading saved Boruta results for dataset:", dataset_name))
  load(result_file)
  load(flag_file)
  message(paste("Boruta already executed for dataset:", dataset_name))
}

```


```{r}
# Print Boruta summary
print(boruta_result)
```

```{r}
# Get selected features
final_features <- getSelectedAttributes(boruta_result, withTentative = FALSE)
final_features
```


```{r}
# Plot Boruta results
plot(boruta_result, xlab = "", las = 2, 
     cex.main = 0.8, cex.axis = 0.6, 
     main = paste("Boruta Feature Selection"))
```

```{r}
print(paste('The number of selected variables are', length(final_features)))
```

## Create a Data frame with the Selected Features

```{r}
# Subset the training and test data using the selected features
X_train_selected <- X_train[, final_features, drop = FALSE]  
X_test_selected <- X_test[, final_features, drop = FALSE]   

# Keep y_train and y_test as-is 
y_train_selected <- y_train
y_test_selected <- y_test


#Merge data_frame
train_merged <- cbind(X_train_selected, binary_class = y_train_selected)
test_merged <- cbind(X_test_selected, binary_class = y_test_selected)

dim(train_merged)
dim(test_merged)
table(train_merged$binary_class)
table(test_merged$binary_class)

```

## Create a Data frame with the Selected Features for SMOTE Data

```{r}
X_train_selected_smote <- X_train_smote[,final_features, drop = FALSE]
X_test_selected_smote <- X_test[, final_features, drop = FALSE] 

# Keep y_train and y_test as-is 
y_train_selected_smote <- y_train
y_test_selected_smote <- y_test

#Merge data_frame
train_merged_smote <- cbind(X_train_selected_smote, binary_class = smote_data$binary_class)
test_merged_smote <- cbind(X_test_selected_smote, binary_class = y_test_selected_smote)

dim(train_merged_smote)
dim(test_merged_smote)
table(train_merged_smote$binary_class)
table(test_merged_smote$binary_class)
```

## Check correlations by HeatMaps

```{r}

n_features <- length(final_features)

# Calculate the correlation matrix 

#Imbalanced Data

#Calculate correlations
corr <- cor(train_merged[, -(n_features+1)])

# Create a heat map
par(cex.main = 0.6)
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("blue", "white", "red"))(50),
        main = "Correlation Heatmap")

# Add legend

legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("blue", "white", "red"))(5), 
       title = "Correlation", cex = 0.8) 

```

```{r}
# SMOTE Data
par(cex.main = 0.6)
corr <- cor(train_merged_smote[, -(n_features+1)])
# Create heat map
heatmap(corr, symm = TRUE, 
        col = colorRampPalette(c("#440154", "#21908C", "gold"))(50),
        main = 'Correlation Heatmap',cex.main = 0.6)  

# Add legend
legend("right", 
       legend = seq(-1, 1, by = 0.5),  # Correlation range
       fill = colorRampPalette(c("#440154", "#21908C", "gold"))(5), 
       title = "Correlation", 
       cex = 0.8) 
```

-   The correlation heat maps reveal that some variables remain highly correlated. To address this issue and mitigate multicollinearity, we will employ Principal Component Analysis (PCA)

## PreProcess by PCA, Centering and Scaling Data

```{r}

# Imbalanced Data
pca_Preprocess <- preProcess(train_merged, method = c("center", "scale", "pca"))
train_mergedPreprocessed <- predict(pca_Preprocess, train_merged) # train data
test_mergedPreprocessed <- predict(pca_Preprocess,test_merged) # apply on test data

# SMOTE Data
pca_Preprocess_smote <- preProcess(train_merged_smote,method = c("center", "scale", "pca"))
train_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,train_merged_smote)
test_merged_smote_Preprocessed <- predict(pca_Preprocess_smote,test_merged_smote)

# Check dimensions of the preprocessed data to ensure consistency

dim(train_mergedPreprocessed)
dim(test_mergedPreprocessed)

dim(train_merged_smote_Preprocessed)
dim(test_merged_smote_Preprocessed)
```

## Train Models

-   We will train models based on the reduced number of features.

```{r}
# Set random seed for reproducibility
set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# One SE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Hyperparameter Tuning
# Set tune grid to find optimal value of size and decay on cross-validation
# Size is the number of number of units (neurons) in the hidden layer of the neural network
# Decay is the regularization parameter; adds penalty to the loss function

grid <- expand.grid(.size=seq(1,10, by =1),.decay=c(0,0.001,0.01,0.1,0.2,0.3,0.4,0.5))
```

```{r cache=TRUE}
#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_nnet_raw<-train(binary_class ~., 
                         data = train_mergedPreprocessed,
                         method = 'nnet',
                         tuneGrid = grid,
                         trControl = control, 
                         metric = 'Accuracy',MaxNWts = 1000, maxit = 300) #unmodified data
#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()
fit_nnet_smote_raw <- train(binary_class ~., 
                         data = train_merged_smote_Preprocessed,
                         method = 'nnet', tuneLength = 10,
                         trControl = control, 
                         tuneGrid = grid,
                         metric = 'Accuracy',MaxNWts = 1000, maxit = 300) # SMOTE data

#End_time
end_time<-Sys.time()

model_training_time<-end_time-start_time

stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```

```{r}
# Display model results for imbalanced data
print(fit_nnet_raw)
```

```{r}
# Display results for data balanced by SMOTE
print(fit_nnet_smote_raw)
```

## Save the models for future use

```{r}
saveRDS(fit_nnet_raw, "fit_nnet_raw.rds")
saveRDS(fit_nnet_smote_raw, "fit_nnet_smote_raw.rds")
```

## Plot the cross-validation results —–> Model Selection 

```{r warning=FALSE}

# Imbalanced Data

# Get the best accuracy based on best size and decay
best_size <- fit_nnet_raw$bestTune$size
best_decay <- fit_nnet_raw$bestTune$decay
best_accuracy <- fit_nnet_raw$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_raw, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

```{r warning=FALSE}
# SMOTE Data
# Get the best accuracy based on best size and decay
best_size <- fit_nnet_smote_raw$bestTune$size
best_decay <- fit_nnet_smote_raw$bestTune$decay
best_accuracy <- fit_nnet_smote_raw$results %>%
  filter(size == best_size & decay == best_decay) %>%
  pull(Accuracy)

# Print the best accuracy
ggplot(fit_nnet_smote_raw, na.rm = TRUE) +
  geom_vline(
    xintercept = best_size, 
    color = "blue", 
    linetype = "dashed", 
    size = 0.7
  ) +
  geom_hline(
    yintercept = best_accuracy, 
    color = "black", 
    linetype = "dashed", 
    size = 0.7
  ) +
  labs(
    title = "Hyperparameter Tuning for Neural Network Model",
    x = "Size",
    y = "Accuracy"
  ) +
  theme_solarized_2() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 9),
    legend.title = element_text(size = 8)
  )
```

## Display the cross-validation results

```{r}
# Select the best results row 
selected_row <- fit_nnet_raw$results %>% 
  filter(size==fit_nnet_raw$bestTune$size & 
           decay ==fit_nnet_raw$bestTune$decay)


kable(selected_row) # unmodified data 
```

```{r}
# Select the best results row 
selected_row_smote <- fit_nnet_smote_raw$results %>% 
  filter(size==fit_nnet_smote_raw$bestTune$size & 
           decay ==fit_nnet_raw$bestTune$decay) # SMOTE data 

kable(selected_row_smote)
```

## Display CV Results in Tables

```{r}
kable(fit_nnet_raw$results)
```

```{r}
kable(fit_nnet_smote_raw$results)
```

```{r}
kable(selected_row)
```


## Confusion matrix for the training data

```{r}
confusionMatrix(fit_nnet_raw$pred$pred,
                fit_nnet_raw$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_nnet_smote_raw$pred$pred,
fit_nnet_smote_raw$pred$obs) # SMOTE data
```
## Visualize the NNET Model Architecture

```{r}
library(NeuralNetTools)
# Unmodified data
plotnet(fit_nnet_raw$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
 
```

```{r}
#SMOTE data
plotnet(fit_nnet_smote_raw$finalModel,
        y_names = 'Class',circle_col = c("lightblue", "pink", 
        "gold", "green","maroon"),
        alpha = 0.7,pos_col = "darkgreen",
        neg_col = "red",cex_val = 0.8,cex_text = 0.7,bord_col = "gray50")
```


## Test the model performance on test data for unbalanced trained data

```{r}
# Predict on the test set 

test_nnet_unmodified <- predict(fit_nnet_raw,newdata=test_mergedPreprocessed)

# Get the confusion matrix
test_nnet_unmodified_CMtrx <- confusionMatrix(test_nnet_unmodified,
                                             test_mergedPreprocessed$binary_class)
test_nnet_unmodified_CMtrx
```

```{r}
kable(test_nnet_unmodified_CMtrx$byClass)
```


```{r}
#View the results as knitr table
kable(test_nnet_unmodified_CMtrx$table)
```


```{r}
# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_nnet <- test_nnet_unmodified_CMtrx$table
TP <- cf_table_nnet[1,1]
TN <- cf_table_nnet[2,2] 
FP <- cf_table_nnet[2,1] 
FN <- cf_table_nnet[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model performance on test data using SMOTE trained model

```{r}
# Predict on the test set 

test_nnet_smote <- predict(fit_nnet_smote_raw,
                          newdata=test_merged_smote_Preprocessed)

# Get the confusion matrix
test_nnet_smote_CMtrx <- confusionMatrix(test_nnet_smote,
                                             test_merged_smote_Preprocessed$binary_class)
test_nnet_smote_CMtrx
```

```{r}
kable(test_nnet_smote_CMtrx$byClass)
```

```{r}
# Calculate the Matthews Correlation coefficient: SMOTE Data

# Extracting the components of the confusion matrix
cf_table_nnet_smote <- test_nnet_smote_CMtrx$table
TP <- cf_table_nnet_smote[1,1]
TN <- cf_table_nnet_smote[2,2] 
FP <- cf_table_nnet_smote[2,1] 
FN <- cf_table_nnet_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

```{r}
# Compare model performance
results <- resamples(list(nnet = fit_nnet_raw, nnet_smote = fit_nnet_smote_raw))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert re-sampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Performance Metrics")
```

```{r}
# Boxplot for accuracy comparison
bwplot(results, col = 'red')
```

```{r}
# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics
```

```{r}
p1<-ggplot(data = as.data.frame(test_nnet_unmodified_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix nnet Unbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(data = as.data.frame(test_nnet_smote_CMtrx$table), aes(x = Reference, y = Prediction)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label = sprintf("%0.0f", Freq)), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Actual Class", y = "Predicted Class", color = 'black',title = 'Confusion Matrix nnet with SMOTE')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}

#SMOTE data

# Generate the ROC curve
roc_curve <- roc(test_merged_smote_Preprocessed$binary_class,
                 as.numeric(test_nnet_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("nnet"), col = 'blue', lwd = 3)
```

## Conclusion

- The neural network based on Boruta feature selection on raw spectra data shows a relatively lower performance compared to models built with preprocessed spectra. 