---
title: "Detection of Coffee Adulteration Using Hyperspectral Imaging and Support Vectors Machine (SVM): MSC+SG+1D Spectra"
freeze: false
author: "Derick Malavi"
format: 
  html:
    embed-resources: true   
    toc: true
    toc-title: "Table of Contents"
    toc-depth: 3
    css: css_style_3.txt
    theme: 'sky'
    geometry: "left=2cm, right=2cm, top=2cm, bottom=2cm"
editor_options: 
  chunk_output_type: inline
  output: true
---

## Introduction

This project aims to distinguish authentic coffee from adulterated samples using Support Vector Machines (SVM), a powerful machine learning algorithm known for its effectiveness in handling complex classification tasks. By analyzing spectral data, the model will identify key features that differentiate pure coffee from blends with varying levels of adulteration, ensuring high accuracy and reliability in detection.

```{r}
# Load libraries
warning = FALSE  
  suppressWarnings(suppressMessages({    
  library(caret) 
  library(ggplot2)    
  library(dplyr)   
  library(readxl)   
  library(readr)    
  library(pls)   
  library(janitor)  
  library(class)   
  library(MLmetrics) 
  library(MLeval)  
  library(themis)  
  library(ROSE) 
  library(parallel)    
  library(doParallel)    
  library(foreach)    
  library(pheatmap)   
  library(mdatools)    
  library(nnet)
  library(e1071) 
  library(FactoMineR)
  library(doSNOW)
  library(MASS)
  library(pROC)
  library(factoextra)
  library(gridExtra)
  library(kernlab)}))
```

## Load and inspect data: msc_sg_1d spectra

```{r}
hsi_msc_sg_1d <- read.csv("coffee_instant_msc_sg_1d.csv")

# Ensure we have the same number of dimensions and the range of our columns/covariates
dim(hsi_msc_sg_1d)

# Check a few column names
col_check <- colnames(hsi_msc_sg_1d[,c(1:6)])
col_check
# The spectral values start from column # 6-------> 229

# Convert the class to a factor
hsi_msc_sg_1d$binary_class <- as.factor(hsi_msc_sg_1d$binary_class)
```

```{r}
# Get the X and y variables
X <- hsi_msc_sg_1d[,c(6:229)]
y <- hsi_msc_sg_1d[,2]
```

## Visualize X

```{r}
# Create histogram 
hist(as.matrix(X), 
     col = "lightblue",
     border = "white",  
     main = "Distribution of Spectral Intensity Values",  
     xlab = "Spectral Intensity (AU)",  
     ylab = "Frequency",  
     las = 1,  
     cex.main = 1.2,  
     cex.lab = 1.1,  
     breaks = 30)  

# Add grid lines 
grid(nx = NA, ny = NULL, col = "gray90", lty = "dotted")

# Add box around plot 
box(lwd = 2)  
```


## Split samples into training and test sets

```{r}
# Subset samples for cross-validation and external validation (70/30)
# Record 1 under cal_val variable is for calibration while 2 will be for testing the model

sample<-hsi_msc_sg_1d$cal_val== 1 

# Split to train and test sets
X_train <- X[sample,]
X_test <- X[!sample,]
y_train <-y[sample]
y_test <- y[!sample]

# Check length of the subsets for training and validation
dim(X_train)
dim(X_test)
length(y_train)
length(y_test)

# Merge the test and train data
train_data <- data.frame(X_train, binary_class = y_train)
test_data <- data.frame(X_test, binary_class = y_test)

# Check composition of the y variable for both train and test sets
table(train_data$binary_class)
table(test_data$binary_class)
```

```{r}
# Create a bar plot for class distribution in train_data
train_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

## Balance the data using SMOTE

```{r}
# Due to data imbalance, we will use SMOTE to balance the data set

library(DMwR)

# Combine the dataset
smote_data <- SMOTE(binary_class ~ ., data = train_data,
                    perc.over = 1900, perc.under = 100)

# Check new class distribution
table(smote_data$binary_class)

# Extract the over sampled data
X_train_smote <- smote_data[, -ncol(smote_data)]  
y_train_smote <- smote_data$binary_class    
```

```{r}
# Visualize the balanced data 

smote_data %>%
  count(binary_class) %>%
  ggplot(aes(x = binary_class, y = n, fill = binary_class)) +
  geom_bar(stat = "identity") +
  labs(title = "Class Distribution in SMOTE Train Data",
       x = "Class",
       y = "Count") +
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5))
```

# Reduction Dimensional (PCA), Centering and Scaling Data

```{r}
# Reduction dimensionality to eliminate collinearity, centering and scaling

# Imabalanced Data
pca_Preprocess<- preProcess(train_data, method = c("center", "scale", "pca"))
data_trainPreprocessed <- predict(pca_Preprocess, train_data)
data_testPreprocessed <- predict(pca_Preprocess, test_data)

#SMOTE data
pca_Preprocess_smote<- preProcess(smote_data, method = c("center", "scale", "pca"))
data_trainPreprocessed_smote <- predict(pca_Preprocess_smote, smote_data)
data_testPreprocessed_smote <- predict(pca_Preprocess_smote, test_data)
```

## Set Up Training Parameters

```{r}
# Set random seed for reproducibility

set.seed (123) 

# Set up the training control(10 folds 10 times cross_validation)
control <- trainControl(method = "repeatedcv", number = 10, repeats = 10, 
                        classProbs = TRUE, savePredictions = "final", 
                        summaryFunction = multiClassSummary,
                        selectionFunction = 'oneSE', allowParallel = TRUE)

# OneSE rule will be used to select simplest yet optimal model within the 'best model'
```

```{r}
#Set the metric as accuracy 
metric<-"Accuracy"
```

```{r}
# Detect the number of cores
num_cores <- detectCores()
# Print the number of cores
print(paste("Number of cores available:", num_cores))
```

```{r}
# Hyperparameter Tuning
# We will use tuneLength to automatically determine the optimal parameters 
```

## Train the Models

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

# Start_time
start_time<-Sys.time()

#Train the model using cross-validation (10 folds, 10 times)
fit_svm_msc_sg_1d<-train(binary_class~., data = data_trainPreprocessed, 
                      method = 'svmRadial', tuneLength = 10, 
                      trControl = control, metric = 'Accuracy') #unmodified data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)
```


```{r}
print(fit_svm_msc_sg_1d)
```

```{r cache=TRUE}

#Register cluster for caret to train the models in parallel
set.seed(123)
cl<-makeCluster(6,type = "SOCK")
suppressWarnings(suppressMessages(
  registerDoSNOW(cl)))

fit_svm_smote_msc_sg_1d<-train(binary_class~., data = data_trainPreprocessed_smote,
                            method ='svmRadial',tuneLength = 10, 
                            trControl = control, metric = 'Accuracy')# SMOTE data
#End_time
end_time<-Sys.time()
model_training_time<-end_time-start_time
stopCluster(cl) # Stop the parallel run cluster

print(model_training_time)

```

```{r}
print(fit_svm_smote_msc_sg_1d)
```

## Save the models for future use

```{r}
saveRDS(fit_svm_msc_sg_1d, "fit_svm_msc_sg_1d.rds")
saveRDS(fit_svm_smote_msc_sg_1d, "fit_svm_smote_msc_sg_1d.rds")
```

## Plot the cross-validation results ----> Model selection

```{r}
# Imbalanced Data
fit_svm_msc_sg_1d
C_data<-fit_svm_msc_sg_1d$results
best_C = fit_svm_msc_sg_1d$bestTune$C
y_Acc = C_data$Accuracy[C_data$C == best_C]
plot(C_data$C,C_data$Accuracy, type = "b",col = "red",pch = 20,cex.axis = 0.8, 
     xlab = "Cost", ylab = "Accuracy (%)",cex.lab = 0.9)
# Add line from x-axis to the specific point
segments(x0 = best_C, y0 = 0, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)
```

```{r}
# Balanced Data

fit_svm_smote_msc_sg_1d
C_data<-fit_svm_smote_msc_sg_1d$results
best_C = fit_svm_smote_msc_sg_1d$bestTune$C
y_Acc = C_data$Accuracy[C_data$C == best_C]
plot(C_data$C,C_data$Accuracy, type = "b",col = "blue",pch = 20,cex.axis = 0.8, 
     xlab = "Cost", ylab = "Accuracy (%)",cex.lab = 0.9,ylim = c(0,1))
# Add line from x-axis to the specific point
segments(x0 = best_C, y0 = 0, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)

# Add line from y-axis to the specific point
segments(x0 = 0, y0 = y_Acc, x1 = best_C, y1 = y_Acc, col = "black", lty = 2)
```

## Display the cross-validation results

```{r}
print(paste('The optimal parametrers for training the SVM model are sigma value of', round(as.numeric(fit_svm_msc_sg_1d$bestTune$sigma),2), 'and Cost parameter of', fit_svm_msc_sg_1d$bestTune$C)) # unmodified data 

print(paste('The optimal parametrers for training the SVM model are sigma value of', round(as.numeric(fit_svm_smote_msc_sg_1d$bestTune$sigma),2), 'and Cost parameter of', fit_svm_smote_msc_sg_1d$bestTune$C)) # SMOTE data 
```



```{r}
# Imbalanced data
knitr::kable(fit_svm_msc_sg_1d$results)
```

```{r}
# Select the best results row 
selected_row <- fit_svm_msc_sg_1d$results %>% filter(C==as.numeric(fit_svm_msc_sg_1d$bestTune$C))

knitr::kable(selected_row)
```

```{r}
# SMOTE Data
knitr::kable(fit_svm_smote_msc_sg_1d$results)

```

```{r}
# Select the best results row 
selected_row_smote <- fit_svm_smote_msc_sg_1d$results %>% filter(C==as.numeric(fit_svm_smote_msc_sg_1d$bestTune$C))

knitr::kable(selected_row_smote)
```

## Confusion matrix for the training data

```{r}
confusionMatrix(fit_svm_msc_sg_1d$pred$pred,fit_svm_msc_sg_1d$pred$obs) # unmodified data
```

```{r}
confusionMatrix(fit_svm_smote_msc_sg_1d$pred$pred,fit_svm_smote_msc_sg_1d$pred$obs) # SMOTE data
```

## Test the model pesvmormance on test data for unbalanced trained data

```{r}
# Predict on the test set 
set.seed(123)
test_svm_unmodified <- predict(fit_svm_msc_sg_1d,newdata=data_testPreprocessed)

# Get the confusion matrix
test_svm_unmodified_CMtrx <- confusionMatrix(test_svm_unmodified,
                                             data_testPreprocessed$binary_class)
test_svm_unmodified_CMtrx

knitr::kable(test_svm_unmodified_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_svm_unmodified_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm <- test_svm_unmodified_CMtrx$table
TP <- cf_table_svm[1,1]
TN <- cf_table_svm[2,2] 
FP <- cf_table_svm[2,1] 
FN <- cf_table_svm[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Test the model pesvmormance on test data using SMOTE trained model

```{r}
# Predict on the test set 
set.seed(123)
test_svm_smote <- predict(fit_svm_smote_msc_sg_1d,newdata=data_testPreprocessed_smote)

# Get the confusion matrix
test_svm_smote_CMtrx <- confusionMatrix(test_svm_smote,
                                             data_testPreprocessed_smote$binary_class)
test_svm_smote_CMtrx

knitr::kable(test_svm_smote_CMtrx$byClass)

#View the results as knitr table
knitr::kable(test_svm_smote_CMtrx$table)

# Calculate the Matthews Correlation coefficient 

# Extracting the components of the confusion matrix
cf_table_svm_smote <- test_svm_smote_CMtrx$table
TP <- cf_table_svm_smote[1,1]
TN <- cf_table_svm_smote[2,2] 
FP <- cf_table_svm_smote[2,1] 
FN <- cf_table_svm_smote[1,2]

# Calculating MCC
MCC <- (TP * TN - FP * FN) / sqrt((TP + FP) * (TP + FN) * (TN + FP) * (TN + FN))

# Printing the MCC value
print(paste("The MCC value is for the model",MCC))
```

## Compare the models for unbalanced and balanced data

```{r}
library(knitr)
# Compare model pesvmormance
results <- resamples(list(svm = fit_svm_msc_sg_1d, svm_smote = fit_svm_smote_msc_sg_1d))

# Print summary of re-sampling results
resample_summary <- summary(results)$statistics

# Convert resampling statistics to a data frame
resample_df <- do.call(rbind, lapply(resample_summary, function(x) x[, "Mean", drop = FALSE]))

# Create a neat table
kable(resample_df, caption = "Model Pesvmormance Metrics")

# Boxplot for accuracy comparison
bwplot(results)

# Pairwise comparisons
model_differences <- diff(results)
model_differences$statistics

```

## Visualize the confusion matrix tables

```{r}
# Imbalanced data
plt_Imb <- as.data.frame(test_svm_unmodified_CMtrx$table)
plt_Imb$Reference <- factor(plt_Imb$Reference, levels=rev(levels(plt_Imb$Reference)))

# SMOTE data
plt_smote <- as.data.frame(test_svm_smote_CMtrx$table)
plt_smote$Reference <- factor(plt_smote$Reference, levels=rev(levels(plt_smote$Reference)))


p1<-ggplot(plt_Imb, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "gray84", high = "darkorange3", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: svm Imbalanced Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))


p2<-ggplot(plt_smote, aes(y = Reference, x = Prediction, fill=Freq)) +
  geom_tile(aes(fill = Freq), color = "black") +
  geom_text(aes(label=Freq), vjust = 1,color = "black",size = 4) +
  scale_fill_gradient(low = "white", high = "#99ccff", name = "Frequency") +
  theme_minimal() +
  labs(x = "Predicted Class", y = "Actual Class", color = 'black',title = 'Confusion Matrix: svm SMOTE Data')+
  theme(
    legend.position =  "none",
    axis.text.y = element_text(color = "black",size = 8),
    axis.text.x = element_text(color = "black",size = 8),
    axis.title.x  = element_text(size = 8),
    axis.title.y = element_text(size = 8),aspect.ratio = 1,
    plot.title = element_text(size = 8,hjust = 0.5))

grid.arrange(p1,p2, nrow =1)
```

## ROC Curve

```{r}
#SMOTE data
# Generate the ROC curve
roc_curve <- roc(data_testPreprocessed_smote$binary_class,
                 as.numeric(test_svm_smote),
                 levels = c("adulterated", "pure_arabica"),
                 positive = 'adulterated')
```

```{r}
# Calculate the AUC
auc_value <- auc(roc_curve)

# Plot the ROC curve
plot(roc_curve, col = 'blue', lwd = 2,
     main = paste("ROC Curve (AUC =", round(auc_value, 2), ")"),
     xlab = "False Positive Rate (1 - Specificity)",
     ylab = "True Positive Rate (Sensitivity)",
     xlim = c(0, 1), ylim = c(0, 1))  
abline(a = 0, b = 1, lty = 2, col = "black")
grid(nx = NULL, ny = NULL, col = "lightgray", lty = "dotted")  
legend("bottomright", legend = c("svm"), col = 'blue', lwd = 3)
```

### Misclassifications

```{r}
## Imbalanced
observed <- y_test
predicted <- test_svm_unmodified
perc <- hsi_msc_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

imbal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_imbal <- imbal_class %>% filter(observed!=predicted)
misclas_imbal
```

```{r}
table(misclas_imbal$perc_adulter)
```

```{r}
# SMOTE Data
observed <- y_test
predicted <- test_svm_smote
perc <- hsi_msc_sg_1d %>% filter(cal_val==2) %>% dplyr::select(perc_adulter)

bal_class <- as.data.frame(cbind(observed,predicted,perc))
misclas_smote <- bal_class %>% filter(observed!=predicted)
misclas_smote
```

```{r}
table(misclas_smote$perc_adulter)
```

